---
title: "LadnerCommunityChanges"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#citation("ggplot2")

```

```{r libraries}
library(tidyverse)
library(vegan)
library(codyn)
library(dplyr)
library(indicspecies) # indicator species analysis
library(ggplot2)
library(dendextend) # dendrogram formatting
library(ggpubr) # plot panel format
library(ape)
library(kableExtra) # for making tables
library(RColorBrewer)

#ordination ellipses, see http://kembellab.ca/r-workshop/biodivR/SK_Biodiversity_R.html
# approx halfway down the page
```

```{r Read data & standardize names, echo=FALSE}

d2019 <- read.csv("d2019_formatted.csv")
# 2019_Lane_BINNED-NRCS.csv
#view(d2019)

d99 <- read.csv("d1999_formatted.csv") 

oldnames99 = c("Agrostis_alba","Aster_eatonii", "Mentha_citrata", "Scirpus_validus", "Trifolium_wormskjoldii")
newnames99 = c("Agrostis_stolonifera","Symphotrichum_subspicatum", "Mentha_aquatica", "Schoenoplectus_tabernaemontani", "Trifolium_wormskioldii")

d1999 <- d99 %>% 
  rename_at(vars(oldnames99), ~ newnames99) %>% 
  mutate(Year = 1999)  %>% 
  relocate(Year)


#1999_Denoth-NRCS.csv
# if first column has a byte order mark
#colnames(d1999)[1] <- gsub('^...','',colnames(d1999)[1])
#view(d1999)

d79 <- read.csv("d1979_formatted.csv") 

oldnames79 = c("Agrostis_alba","Aster_eatonii", "Galium_cymosum", "Mentha_citrata", "Hypericum_formosum", "Puccinella_pauciflora", "Scirpus_validus", "Trifolium_wormskjoldii")
newnames79 = c("Agrostis_stolonifera","Symphotrichum_subspicatum", "Galium_trifidum", "Hypericum_scouleri", "Mentha_aquatica", "Puccinellia_pauciflora", "Schoenoplectus_tabernaemontani", "Trifolium_wormskioldii")

d1979 <- d79 %>% 
  rename_at(vars(oldnames79), ~ newnames79) %>% 
  mutate(Year = 1979) %>% 
  relocate(Year)
# 1979_Bradfield-NRCS.csv
# if first column has a byte order mark
#colnames(d1979)[1] <- gsub('^...','',colnames(d1979)[1])
#view(d1979)



```

```{r align data across time}

#filter only plots that fall along the same lengths of transect sampled in all years 
align19 <- filter(d2019, Quadrat > 0)
#nrow(align19) #74

align99 <- d1999 %>% filter((between(Quadrat, 9,16)) | (between(Quadrat, 20,32)) | (between(Quadrat, 37,44)) | (between(Quadrat, 46,50)) | (between(Quadrat, 55,69)) | (between(Quadrat, 72,88)) | (between(Quadrat, 93,108)))
#nrow(align99) #82

align79 <- d1979 %>% filter((between(Quadrat, 9,16)) | (between(Quadrat, 20,32)) | (between(Quadrat, 37,44)) | (between(Quadrat, 46,50)) | (between(Quadrat, 55,69)) | (between(Quadrat, 72,88)) | (between(Quadrat, 93,108)))
#nrow(align79) #82


```



RQ 1: "Are assemblages still characterized by the same dominant species?
```{r Cluster analysis - Euclidean}

###########################
# 2019
###########################

d2019_select <- select(align19, -(Year:Quadrat))
ncol(d2019_select) #43 species

#name leaves
d2019_unite <- unite(align19, plot, Quadrat, remove = TRUE, na.rm = FALSE)
#d2019_unite_all <- unite(d2019, plot, Quadrat, remove = TRUE, na.rm = FALSE)

rownames(d2019_select) <- d2019_unite$plot

#cluster - method ward.D does NOT square the dissimilarities. 
d2019_ward <- d2019_select %>% 
  dist(method = "euclidean") %>% 
  hclust(method = "ward.D") %>% 
  as.dendrogram()

#library(dendextend)
mycols2019 <- c("#0072B2", "#E69F00", "#009E73")
dend2019 <-  as.dendrogram(d2019_ward) %>%
   set("branches_lwd", 1) %>% # Branches line width
   set("branches_k_color", mycols2019, k = 3) %>% # Color branches by groups
   set("labels_col", "white") # mask labels by changing color
   #set("labels_colors", mycols, k = 3) %>%  # Color labels by groups
   #set("labels_cex", 0.5) # Change label size
#fviz_dend(dend2) 

#changing font sizes/formatting: https://stackoverflow.com/questions/17232178/change-label-size-of-cluster-dendrogram-in-r-3-01
par(cex=2.5, mar=c(2.5,4,2,1))
plot(dend2019, ylab = "Euclidean Distance", main = "2019")

###########################
# 1999
###########################

d1999_select <- select(align99, -(Year:Quadrat))
ncol(d1999_select) #47

#name leaves
d1999_unite <- unite(align99, plot, Quadrat, remove = TRUE, na.rm = FALSE)

rownames(d1999_select) <- d1999_unite$plot

#cluster - method ward.D does NOT square the dissimilarities. 
d1999_ward <- d1999_select %>% 
  dist(method = "euclidean") %>% 
  hclust(method = "ward.D") %>% 
  as.dendrogram()

#library(dendextend)
mycols1999 <- c("#009E73", "#0072B2", "#E69F00")
dend1999 <-  as.dendrogram(d1999_ward) %>%
   set("branches_lwd", 1) %>% # Branches line width
   set("branches_k_color", mycols1999, k = 3) %>% # Color branches by groups
   set("labels_col", "white") # mask labels by changing color
   #set("labels_colors", mycols, k = 3) %>%  # Color labels by groups
   #set("labels_cex", 0.5) # Change label size
#fviz_dend(dend2) 

#changing font sizes/formatting: https://stackoverflow.com/questions/17232178/change-label-size-of-cluster-dendrogram-in-r-3-01
par(cex=2.5, mar=c(2.5,4,2,1))
plot(dend1999, ylab = "Euclidean Distance", main = "1999")

###########################
# 1979
###########################

d1979_select <- select(align79, -(Year:Quadrat))
#ncol(d1979_select) #55
#view(d1979_select)

#name leaves
d1979_unite <- unite(align79, plot, Quadrat, remove = TRUE, na.rm = FALSE)

rownames(d1979_select) <- d1979_unite$plot


d1979_ward <- d1979_select %>% 
  dist(method = "euclidean") %>% 
  hclust(method = "ward.D") %>% 
  as.dendrogram()

#library(dendextend)
mycols1979 <- c("#009E73", "#0072B2", "#E69F00")
dend1979 <-  as.dendrogram(d1979_ward) %>%
   set("branches_lwd", 1) %>% # Branches line width
   set("branches_k_color", mycols1979, k = 3) %>% # Color branches by groups
    set("labels_col", "white") # mask labels by changing color
   #set("labels_colors", mycols, k = 3) %>%  # Color labels by groups
   #set("labels_cex", 0.5) # Change label size
#fviz_dend(dend2) 

#changing font sizes/formatting: https://stackoverflow.com/questions/17232178/change-label-size-of-cluster-dendrogram-in-r-3-01
par(cex=2.5, mar=c(2.5,4,2,1))
plot(dend1979, ylab = "Euclidean Distance", main = "1979")

###########################
# MULTI-PANEL DENDROGRAM
###########################


#install.packages("svglite")
# library(svglite)
# 
# svglite("Fig3.svg", width=8, height=5)
# par(mfrow = c(1,3), mar = c(6, 5, 5, 3), xpd = TRUE, mgp=c(1.75, .75,  0))
# plot(dend1979, main = "1979")
# mtext(text = "Euclidean Distance", side = 2, line = 3, col = "black")
# plot(dend1999, main = "1999")
# legend("bottom", c("Sedge", "Fescue", "Bogbean"),
#         title = "Assemblage Type", horiz = TRUE, inset = c(0, -0.2),
#         col = c("#009E73", "#E69F00", "#0072B2"), pch = rep(15,2),
#         bty = "n", pt.cex = 1.5, text.col = "black")
# plot(dend2019, main = "2019")
# dev.off()

```

```{r indicator species analysis - Euclidean}
## see cran vignette for details https://cran.r-project.org/web/packages/indicspecies/vignettes/IndicatorSpeciesAnalysis.html#indicator-species-analysis-using-multipatt 

#following https://jkzorz.github.io/2019/07/02/Indicator-species-analysis.html 

###########################
# 2019
###########################

#Note: func "IndVal.g" incorporates a correction for unequal group sizes, while "IndVal" does not (p. 10)
#Note: it was tempting to rename the community types by a plant name, but the IndVal requires a numerical matrix

d2019_clstrs <- align19 %>% mutate(Association_Type2 = case_when(Quadrat %in% c(26, 72, 23, 22, 5, 24, 39, 73, 20, 21, 25, 6, 7, 36, 37, 35, 38, 4, 34, 33, 71, 63, 11, 62, 48, 49, 27, 42) ~3, Quadrat %in% c(44, 43, 47, 12, 32, 1, 66, 17, 29, 46, 31, 69, 28, 65, 67, 68, 30, 59) ~2, Quadrat %in% c(19, 40, 74, 8, 10, 13, 9, 54, 50, 15, 55, 41, 70, 2, 52, 57, 53, 56, 64, 3, 16, 51, 58, 14, 18, 45, 60, 61) ~1))
#View(d2019_clstrs)
#Note: new column must be Association_Type2 to join with other data sets for NMS

#move new column to beginning of df (if left at end, new column Association_Type gets counted in the indicspecies)
d2019_grps <- d2019_clstrs %>% 
  select(Association_Type2, everything())
#view(d2019_grps)
nrow(d2019_grps) #74

abund = d2019_grps[,5:ncol(d2019_grps)]
assoc = d2019_grps$Association_Type2

## turn comments off to run - takes a minute
# inv_r.g = multipatt(abund, assoc, func = "r.g", control = how(nperm=9999))
# summary(inv_r.g, indvalcomp = TRUE)

## 1 = Sedge
## 2 = Fescue
## 3 = Bogbean

###########################
# 1999 
###########################

d1999_clstrs <- align99 %>% mutate(Association_Type2 = case_when(Quadrat %in% c(39, 40, 31, 29, 30, 27, 26, 28, 32, 69, 22, 46, 23, 24, 87, 88, 94, 73, 72, 108, 85, 42, 44, 41, 43, 96, 95, 97, 12, 100, 101) ~1, Quadrat %in% c(82, 83, 13, 14, 84, 77, 78, 102, 103, 105, 98, 99, 81, 80, 104, 16, 15, 79) ~3, Quadrat %in% c(10, 11, 37, 38, 20, 93, 107, 9, 106, 62, 25, 59, 61, 56, 57, 58, 68, 60, 67, 55, 86, 65, 66, 63, 64, 76, 74, 75, 50, 47, 49, 21, 48) ~2))

#view(d1999_clstrs)

#move new column to beginning of df (if left at end, new column Association_Type gets counted in the indicspecies)
d1999_grps <- d1999_clstrs %>% 
  select(Association_Type2, everything())
#view(d1999_grps)

abund99 = d1999_grps[,5:ncol(d1999_grps)]
assoc99 = d1999_grps$Association_Type2

## turn comments off to run - takes a minute
# inv99_r.g = multipatt(abund99, assoc99, func = "r.g", control = how(nperm=9999))
# summary(inv99_r.g)

## 1 = Sedge
## 2 = Fescue
## 3 = Bogbean

###########################
# 1979 
###########################

d1979_clstrs <- align79 %>% mutate(Association_Type2 = case_when(Quadrat %in% c(30, 31, 27, 24, 28, 29, 32, 108, 25, 26, 60, 48, 57, 58, 59, 49, 68, 69, 22, 23, 86, 93, 87, 88, 97, 94, 95, 96, 20, 72, 47, 73, 55, 56) ~1, Quadrat %in% c(80, 81, 64, 62, 65, 82, 83, 98, 84, 85, 78, 79, 63, 66, 50, 99, 100, 61, 67) ~3, Quadrat %in% c(38, 74, 11, 12, 107, 9, 10, 76, 106, 102, 101, 103, 42, 43, 14, 15, 40, 41, 39, 44, 21, 37, 16, 13, 104, 105, 75, 46, 77) ~2))
#view(d1979_clstrs)

#move new column to beginning of df (if left at end, new column Association_Type gets counted in the indicspecies)
d1979_grps <- d1979_clstrs %>% 
  select(Association_Type2, everything())
#view(d1979_grps)

abund79 = d1979_grps[,7:ncol(d1979_grps)]
assoc79 = d1979_grps$Association_Type2

## turn comments off to run - takes a minute
# inv79_r.g = multipatt(abund79, assoc79, func = "r.g", control = how(nperm=9999))
# summary(inv79_r.g)

## 1 = Sedge
## 2 = Fescue
## 3 = Bogbean

```

RQ 1 supplemental - using Bray Curtis 
```{r Cluster analysis - Bray-Curtis}
###########################
# 2019
###########################
## Bray-Curtis distances are super tiny in a couple plots if no rounding is specified. This forces the plots with little dissimilarity into their own cluster, overriding k = 3. This can be solved by rounding the distances, as discussed & solved on Stack Overflow https://stackoverflow.com/questions/72282302/specifying-three-cluster-groups-but-dendrogram-clusters-into-four/72298448?noredirect=1#comment127751568_72298448 

#cluster - using object from euclidean chunk
d2019_bray <- d2019_select %>% 
  vegdist(method = "bray") %>% 
  hclust(method = "ward.D") 

d2019_bray$height # view distance values
diff(round(d2019_bray$height,6))

d2019_bray$height <- round(d2019_bray$height, 6) # specify rounding to eliminate errors from "very small" distances
d2019_bray <-  as.dendrogram(d2019_bray)



#hist(d2019_bray)
  
braycol19 <- c("#0072B2", "#009E73", "#E69F00")
dendbray2019 <-  as.dendrogram(d2019_bray) %>%
   set("branches_lwd", 1) %>% # Branches line width
   set("branches_k_color", braycol19, k=3) %>% # Color branches by groups
    set("labels", c()) #remove plot ID labels
   #set("labels_cex", 0.5) # Change label size
plot(dendbray2019, ylab = "Bray-Curtis Distance", main = "2019")



###########################
# 1999
###########################

#name leaves as quadrat
d1999_unite <- unite(align99, plot, Quadrat, remove = TRUE, na.rm = FALSE)
rownames(d1999_select) <- d1999_unite$plot

#cluster - using object from euclidean chunk
d1999_bray <- d1999_select %>% 
  vegdist(method = "bray") %>% 
  hclust(method = "ward.D") %>% 
  as.dendrogram()

braycol99 <- c("#0072B2", "#E69F00", "#009E73")
dendbray1999 <-  as.dendrogram(d1999_bray) %>%
   set("branches_lwd", 1) %>% # Branches line width
   set("branches_k_color", braycol99, k = 3) %>% # Color branches by groups
   set("labels", c()) #remove plot ID labels

plot(dendbray1999, ylab = "Bray-Curtis Distance", main = "1999")


###########################
# 1979
###########################

#name leaves as quadrat
d1979_unite <- unite(align79, plot, Quadrat, remove = TRUE, na.rm = FALSE)
rownames(d1979_select) <- d1979_unite$plot

#cluster - using object from euclidean chunk
d1979_bray <- d1979_select %>% 
  vegdist(method = "bray") %>% 
  hclust(method = "ward.D") %>% 
  as.dendrogram()

braycol79 <- c("#009E73", "#0072B2", "#E69F00")
dendbray1979 <-  as.dendrogram(d1979_bray) %>%
   set("branches_lwd", 1) %>% # Branches line width
   set("branches_k_color", braycol79, k = 3) %>% # Color branches by groups
    set("labels", c()) #remove plot ID labels

plot(dendbray1979, ylab = "Bray-Curtis Distance", main = "1979")


library(svglite)

svglite("Fig_cluster-bray.svg", width=8, height=5)
par(mfrow = c(1,3), mar = c(6, 5, 5, 3), xpd = TRUE, mgp=c(1.75, .75,  0))
plot(dendbray1979, main = "1979")
mtext(text = "Bray-Curtis Distance", side = 2, line = 3, col = "black")
plot(dendbray1999, main = "1999")
legend("bottom", c("Sedge", "Fescue", "Bogbean"),
        title = "Assemblage Type", horiz = TRUE, inset = c(0, -0.2),
        col = c("#009E73", "#E69F00", "#0072B2"), pch = rep(15,2),
        bty = "n", pt.cex = 1.5, text.col = "black")
plot(dendbray2019, main = "2019")
dev.off()


```

```{r indicator species analysis - Bray-Curtis}

###########################
# 2019
###########################

#Note: func "IndVal.g" incorporates a correction for unequal group sizes, while "IndVal" does not (p. 10)
#Note: it was tempting to rename the community types by a plant name, but the IndVal requires a numerical matrix

bray19_clstrs <- align19 %>% mutate(Association_Type2 = case_when(Quadrat %in% c(26, 72, 71, 22, 25, 4, 34, 36, 37, 33, 35, 38, 73, 39, 20, 21, 5, 24, 11, 23, 62, 63, 8, 10, 9, 6, 7, 54, 40, 74, 13, 14) ~"A", Quadrat %in% c(3, 16, 51, 58, 12, 32, 27, 70, 42, 61, 60, 18, 45, 57, 56, 64, 19, 48, 49, 52, 53, 41, 46, 50, 2, 15, 55) ~"B", Quadrat %in% c(59, 65, 67, 28, 31, 67, 28, 31, 69, 68, 17, 30, 29, 43, 44, 1, 47, 66) ~"C"))
#View(bray19_clstrs)

#move new column to beginning of df (if left at end, new column Association_Type gets counted in the indicspecies)
bray19_grps <- bray19_clstrs %>% 
  select(Association_Type2, everything())
#view(d2019_grps)
nrow(bray19_grps) #74

brayabund = bray19_grps[,5:ncol(bray19_grps)]
brayassoc = bray19_grps$Association_Type2

## turn comments off to run - takes a minute
bray_inv = multipatt(brayabund, brayassoc, func = "r.g", control = how(nperm=999))
summary(bray_inv)
# A = METR3/MEAQ
# B = CALY3/MEAR4
# C = PHAR3/FEAR3

###########################
# 1999
###########################

bray99_clstrs <- align99 %>% mutate(Association_Type2 = case_when(Quadrat %in% c(63, 64, 65, 66, 61, 56, 57, 48, 47, 68, 49, 50, 82, 83, 84, 81, 80, 104, 99, 98, 105, 13, 14, 77, 78, 16, 15, 79) ~"A", Quadrat %in% c(10, 11, 93, 107, 9, 106, 37, 38, 74, 75, 76) ~"B", Quadrat %in% c(108, 72, 73, 20, 21, 25, 62, 87, 59, 86, 46, 23, 24, 60, 67, 22, 58, 55, 85, 42, 44, 41, 43, 102, 103, 12, 100, 101, 26, 28, 32, 31, 29, 30, 96, 95, 97, 88, 94, 39, 40, 27, 69) ~"C"))
#View(bray19_clstrs)

#move new column to beginning of df (if left at end, new column Association_Type gets counted in the indicspecies)
bray99_grps <- bray99_clstrs %>% 
  select(Association_Type2, everything())
#view(d2019_grps)
#nrow(bray99_grps) #82

brayabund99 = bray99_grps[,5:ncol(bray99_grps)]
brayassoc99 = bray99_grps$Association_Type2


## turn comments off to run - takes a minute
bray_inv99 = multipatt(brayabund99, brayassoc99, func = "r.g", control = how(nperm=999))
summary(bray_inv99)
# A = METR3/LEOR/MEAQ
# B = FEAR3/PHAR3
# C = CALY3/AGST2


###########################
# 1979
###########################

bray79_clstrs <- align79 %>% mutate(Association_Type2 = case_when(Quadrat %in% c(30, 31, 24, 28, 32, 27, 29, 108, 25, 26, 97, 94, 95, 96, 86, 93, 87, 88, 47, 73, 20, 72, 21, 22, 23, 98, 99, 55, 56, 60, 58, 59, 49, 68, 69, 48, 57) ~"A", Quadrat %in% c(82, 83, 81, 85, 50, 84, 64, 62, 65, 63, 66, 100, 61, 67, 101, 102, 103, 78, 79, 80, 14, 15, 42, 43) ~"B", Quadrat %in% c(11, 9, 10, 76, 106, 107, 38, 74, 37, 40, 41, 39, 44, 16, 104, 105, 46, 77, 75, 12, 13) ~"C"))
#View(bray19_clstrs)

#move new column to beginning of df (if left at end, new column Association_Type gets counted in the indicspecies)
bray79_grps <- bray79_clstrs %>% 
  select(Association_Type2, everything())
#view(d2019_grps)
#nrow(bray99_grps) #82

brayabund79 = bray79_grps[,7:ncol(bray79_grps)]
brayassoc79 = bray79_grps$Association_Type2


## turn comments off to run - takes a minute
bray_inv79 = multipatt(brayabund79, brayassoc79, func = "r.g", control = how(nperm=999))
summary(bray_inv79)
# A = CALY/SALA2/SCTA2
# B = METR3/MYSC/JUAR
# C = FEAR3/SALUL



```


RQ2: "Is total species diversity stable within and between assemblage types over time?"

``` {r join data}
join_40 <- lst(d1979_grps, d1999_grps, d2019_grps) %>% 
  reduce(full_join) %>% 
  replace(is.na(.), 0) %>% 
  select(-Transect) %>% 
  pivot_longer(Agrostis_stolonifera:Scirpus_microcarpus, names_to = "Species", values_to = "Cover_class") %>% 
   mutate(Assemblage = ifelse(Association_Type2 == 1, "Sedge",
                      ifelse(Association_Type2 == 2, "Fescue", "Bogbean"))) %>% 
  select(-Association_Type2) %>% 
  group_by(Year, Assemblage, Species) %>% 
  summarize(mean_cover = mean(Cover_class, na.rm = TRUE))


#View(join_40)

# extra attempt - want a 'total' option by year for mean_cover
join_41 <- lst(d1979_grps, d1999_grps, d2019_grps) %>% 
  reduce(full_join) %>% 
  replace(is.na(.), 0) %>% 
  select(-Transect) %>% 
  pivot_longer(Agrostis_stolonifera:Scirpus_microcarpus, names_to = "Species", values_to = "Cover_class") %>% 
  select(-Association_Type2) %>% 
  group_by(Year, Species) %>% 
  summarize(mean_cover = mean(Cover_class, na.rm = TRUE)) %>% 
   pivot_wider(names_from = Year, values_from = mean_cover) 

```

```{r species abundance table}

# exported table as csv, performed percent change calculations and sorted table in Excel
#unique(join_40$Species)

sp_table <- join_40 %>%
  pivot_wider(names_from = "Year", values_from = "mean_cover") %>%
  filter_at(vars(`1979`, `1999`, `2019`), any_vars(.>0)) %>% 
  mutate(Status = ifelse(Species %in% c("Agrostis_stolonifera", "Alisma_plantago.aquatica", "Cirsium_arvense", "Festuca_arundinacea", "Iris_pseudacorus", "Lycopus_europaeus", "Lythrum_salicaria", "Mentha_aquatica", "Mentha_arvensis", "Myosotis_scorpioides", "Phalaris_arundinacea", "Rumex_conglomeratus"), "Exotic",
                ifelse(Species %in% c("Galium_sp", "Poaceae_misc", "Asteraceae_misc", "Carex_sp1", "Festuca_sp", "Salix_sp"), "Unknown", "Native"))) 

#view(sp_table)

#write.csv(sp_table, "species_table_pct.csv")


sp_table_total <- join_41 %>%
  # pivot_wider(names_from = "Year", values_from = "mean_cover") %>%
  # filter_at(vars(`1979`, `1999`, `2019`), any_vars(.>0)) %>% 
  mutate(Status = ifelse(Species %in% c("Agrostis_stolonifera", "Alisma_plantago.aquatica", "Cirsium_arvense", "Festuca_arundinacea", "Iris_pseudacorus", "Lycopus_europaeus", "Lythrum_salicaria", "Mentha_aquatica", "Mentha_arvensis", "Myosotis_scorpioides", "Phalaris_arundinacea", "Rumex_conglomeratus"), "Exotic",
                ifelse(Species %in% c("Galium_sp", "Poaceae_misc", "Asteraceae_misc", "Carex_sp1", "Festuca_sp", "Salix_sp"), "Unknown", "Native"))) 

#write.csv(sp_table_total, "species_table_total.csv")



```

```{r native/exotic richness over time}

unique(join_40$Species)

endemic <- join_40 %>% 
  mutate(endemic = ifelse(Species %in% c("Asteraceae_misc", "Carex_sp1", "Carex_sp2", "Festuca_sp", "Galium_sp", "Poaceae_misc", "Salix_sp"), "unknown",
                   ifelse(Species %in% c("Agrostis_stolonifera", "Alisma_plantago.aquatica", "Cirsium_arvense", "Festuca_arundinacea", "Impatiens_capensis", "Iris_pseudacorus", "Lycopus_europaeus", "Lythrum_salicaria", "Mentha_aquatica", "Mentha_arvensis", "Myosotis_scorpioides", "Phalaris_arundinacea", "Poa_trivialis", "Polygonum_hydropiper", "Rumex_conglomeratus", "Sonchus_arvensis"), "exotic", "native"))) %>% 
  left_join(join_40) %>% 
  filter_if(is.numeric, all_vars((.) != 0))
  
#View(endemic)

rich_all <- endemic %>% 
        group_by(Year) %>% 
        dplyr::summarize(native_rich = length(endemic[endemic == "native"]), 
            exotic_rich = length(endemic[endemic == "exotic"]), 
            unknown_rich = length(endemic[endemic == "unknown"]), 
            native_cover = mean(mean_cover[endemic == "native"]), 
            exotic_cover = mean(mean_cover[endemic == "exotic"]), 
            unknown_cover = mean(mean_cover[endemic == "unknown"])) %>% 
        mutate(Assemblage = "Total") %>% 
        relocate(Assemblage, .after = Year) %>% 
        replace(is.na(.), 0)
#View(rich_all)

rich_asmbg <- endemic %>% 
        group_by(Year, Assemblage) %>% 
        dplyr::summarize(native_rich = length(endemic[endemic == "native"]), 
            exotic_rich = length(endemic[endemic == "exotic"]), 
            unknown_rich = length(endemic[endemic == "unknown"]), 
            native_cover = mean(mean_cover[endemic == "native"]), 
            exotic_cover = mean(mean_cover[endemic == "exotic"]), 
            unknown_cover = mean(mean_cover[endemic == "unknown"])) %>% 
        replace(is.na(.), 0)
#View(rich_asmbg)

rich <- rich_all %>% 
        full_join(rich_asmbg, by = c("Year", "Assemblage", "native_rich", "exotic_rich", "unknown_rich", "native_cover", "exotic_cover", "unknown_cover")) %>% 
        pivot_longer(native_rich:unknown_rich, names_to = "rich_status", values_to = "richness") %>% 
        pivot_longer(native_cover:unknown_cover, names_to = "cover_status", values_to = "cover")
#View(rich)

rich$rich_status <- factor(rich$rich_status, 
                           levels = c("native_rich", "exotic_rich", "unknown_rich"), 
                           labels = c("Native", "Exotic", "Unknown"))
rich$Year <- factor(rich$Year)

rich_bar <- rich %>% 
            ggplot(aes(x = Year, y = richness, fill = rich_status)) +
            geom_bar(stat="identity", position = position_dodge(width = 1)) +
            theme_classic() + 
            labs(y = "Species Richness", 
                 x = "Year", 
                 fill = "Species Origin") +
            theme(axis.text.x = element_text(size = 12, angle = 30, hjust = 1), strip.text = element_text(size = 12)) +
            facet_grid(. ~ Assemblage) +
            scale_fill_brewer(palette = "Dark2") 
rich_bar

          
rich$cover_status <- factor(rich$cover_status, 
                            levels = c("native_cover", "exotic_cover", "unknown_cover"),
                            labels = c("Native", "Exotic", "Unknown"))
rich$Year <- factor(rich$Year)

cover_bar <- rich %>% 
            ggplot(aes(x = Year, y = cover, fill = cover_status)) +
            geom_bar(stat="identity", 
                     position = position_dodge(width = 1)) +
            theme_classic() + 
            facet_grid(. ~ Assemblage) +
            labs(y = "Mean species cover abundance \n(cover class)", 
                 x = "Year",
                 fill = "Species Origin") +
            theme(text = element_text(size = 12), 
                  panel.spacing.x = unit(1.25, "lines"), 
                  axis.text.x = element_text(angle = 25, hjust = 1)) +
            scale_fill_brewer(palette = "Dark2") 
cover_bar

library(ggpubr)
bar_rich_cover <- ggarrange(rich_bar, cover_bar, ncol = 1, nrow = 2, common.legend = TRUE, legend="bottom")


ggsave("fig-bar-rich-cover.jpg", bar_rich_cover, width = 12, height = 12)

```

```{r diversity calculations}

div.join <- lst(d1979_grps, d1999_grps, d2019_grps) %>% 
  reduce(full_join) %>% 
  replace(is.na(.), 0) %>% 
  mutate(Assemblage = ifelse(Association_Type2 == 1, "Sedge",
                      ifelse(Association_Type2 == 2, "Fescue", "Bogbean"))) %>% 
  relocate(Assemblage, .after = Year) %>% 
  select(-Association_Type2) 
#view(div.join)

# Filter each year; n quadrat = nrow
div79 <- filter(div.join, Year == "1979") %>% 
  select(-Assemblage, -Quadrat, -Year)
nrow(div79) #82
div99 <- filter(div.join, Year == "1999") %>% 
  select(-Assemblage, -Quadrat, -Year)
nrow(div99) #82
div19 <- filter(div.join, Year == "2019")%>% 
  select(-Assemblage, -Quadrat, -Year)
nrow(div19) #74


# total number of species = ncol > 0
sp79 <- div79[, colSums(div79 != 0) > 0]
ncol(sp79) #48

sp99 <- div99[, colSums(div99 != 0) > 0]
ncol(sp99) #45

sp19 <- div19[, colSums(div19 != 0) > 0]
ncol(sp19) #44

##~~~~~~~~~~~~~~~~~~~
# Alpha diversity
##~~~~~~~~~~~~~~~~~~~

sp_plot79 <- apply(sp79 >0,1,sum)
mean(sp_plot79) #9.96
sd(sp_plot79) #3.41

sp_plot99 <- apply(sp99 >0,1,sum)
mean(sp_plot99) #9.55
sd(sp_plot99) #3.30

sp_plot19 <- apply(sp19 >0,1,sum)
mean(sp_plot19) #9.36
sd(sp_plot19) #3.03

##~~~~~~~~~~~~~~~~~~~
# Beta diversity: total n species divided by density
##~~~~~~~~~~~~~~~~~~~

ncol(sp79)/mean(sp_plot79) #4.82
ncol(sp99)/mean(sp_plot99) #4.71
ncol(sp19)/mean(sp_plot19) #4.70


######################################################################
## Diversity metrics by assemblage
######################################################################

#######################
# Sedge 
#######################

# select only plots in sedge group
sedge1979 <- d1979_grps %>% 
  filter(Association_Type2 == 1) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(sedge1979) #34 plots


sedge1999 <- d1999_grps %>% 
  filter(Association_Type2 == 1) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(sedge1999) #31 plots

sedge2019 <- d2019_grps %>% 
  filter(Association_Type2 == 1) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(sedge2019) #28


#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#select only species present
sp_present_sedge1979 <- sedge1979[, colSums(sedge1979 != 0) > 0]
sp_present_sedge1999 <- sedge1999[, colSums(sedge1999 != 0) > 0]
sp_present_sedge2019 <- sedge2019[, colSums(sedge2019 != 0) > 0]

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ncol = # species present
ncol(sp_present_sedge1979) #34
ncol(sp_present_sedge1999) #35
ncol(sp_present_sedge2019) #34

##~~~~~~~~~~~~~~~~~~~
# Alpha diversity
##~~~~~~~~~~~~~~~~~~~

sp_plot_sedge1979 <- apply(sp_present_sedge1979 >0,1,sum)
#[1]   7  7  7 10  9  8  8  5  7  9  6  5 10  6 14  8 11  7 11 11 15  8 11  8 10 12 10  8 10 11  6  7 10  5

sp_plot_sedge1999 <- apply(sp_present_sedge1999 >0,1,sum)
#[1]  9  8  8  5 10  7  6 10  8  9  7  7  7  7 11  9  7 10 13  6  8 12  7  8  7  6 10 11 10  8  5

sp_plot_sedge2019 <- apply(sp_present_sedge2019 >0,1,sum)
#[1]  8 11  9 11  9 11  7  2  5  9  7  9 11  6  8 10 10 12  8  7  3  5 10  5  8

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# species density = mean # sp
mean(sp_plot_sedge1979) #8.74
mean(sp_plot_sedge1999) #8.26
mean(sp_plot_sedge2019) #7.89

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#sd of species density
sd(sp_plot_sedge1979) #2.45
sd(sp_plot_sedge1999) #1.98
sd(sp_plot_sedge2019) #2.69

##~~~~~~~~~~~~~~~~~~~
# Beta diversity: total n species divided by density
##~~~~~~~~~~~~~~~~~~~

ncol(sp_present_sedge1979)/mean(sp_plot_sedge1979) #3.89
ncol(sp_present_sedge1999)/mean(sp_plot_sedge1999) #4.24
ncol(sp_present_sedge2019)/mean(sp_plot_sedge2019) #4.31

#######################
# Fescue 
#######################

fescue1979 <- d1979_grps %>% 
  filter(Association_Type2 == 2) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(fescue1979) #29

fescue1999 <- d1999_grps %>% 
  filter(Association_Type2 == 2) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(fescue1999) #33

fescue2019 <- d2019_grps %>% 
  filter(Association_Type2 == 2) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(fescue2019) #18

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#select only species present
sp_present_fescue1979 <- fescue1979[, colSums(fescue1979 != 0) > 0]
sp_present_fescue1999 <- fescue1999[, colSums(fescue1999 != 0) > 0]
sp_present_fescue2019 <- fescue2019[, colSums(fescue2019 != 0) > 0]

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ncol = # species present
ncol(sp_present_fescue1979) #45
ncol(sp_present_fescue1999) #41
ncol(sp_present_fescue2019) #27

##~~~~~~~~~~~~~~~~~~~
# Alpha diversity
##~~~~~~~~~~~~~~~~~~~

sp_plot_fescue1979 <- apply(sp_present_fescue1979 >0,1,sum)
#[1]   7  7  9 15 13 12 12 17 13  8  7 18 17 11 12 16 18 11  8 11 10 12 14 20 20 17 15 11 11
sp_plot_fescue1999 <- apply(sp_present_fescue1999 >0,1,sum)
#[1] 5  6 10  7  9  6  6  5 17 17 14 13 11 16 12 12  7  8 16 12 14  9 10 10 12 15  5  5  6  8  5  5  7

sp_plot_fescue2019 <- apply(sp_present_fescue2019 >0,1,sum)
#[1] 4 10  7  2  6  7 10 11  7  3  5  8  6  7

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# species density = mean # sp
mean(sp_plot_fescue1979) #10.83
mean(sp_plot_fescue1999) #9.69
mean(sp_plot_fescue2019) #5.83

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#sd of species density
sd(sp_plot_fescue1979) #3.87
sd(sp_plot_fescue1999) #3.96
sd(sp_plot_fescue2019) #2.79

##~~~~~~~~~~~~~~~~~~~
# Beta diversity: total n species divided by density
##~~~~~~~~~~~~~~~~~~~

ncol(sp_present_fescue1979)/mean(sp_plot_fescue1979) #3.66
ncol(sp_present_fescue1999)/mean(sp_plot_fescue1999) #4.23
ncol(sp_present_fescue2019)/mean(sp_plot_fescue2019) #4.63

#######################
# Bogbean 
#######################

bogbean1979 <- d1979_grps %>% 
  filter(Association_Type2 == 3) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(bogbean1979) #19

bogbean1999 <- d1999_grps %>% 
  filter(Association_Type2 == 3) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(bogbean1999) #18

bogbean2019 <- d2019_grps %>% 
  filter(Association_Type2 == 3) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(bogbean2019) #28

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#select only species present
sp_present_bogbean1979 <- bogbean1979[, colSums(bogbean1979 != 0) > 0]
sp_present_bogbean1999 <- bogbean1999[, colSums(bogbean1999 != 0) > 0]
sp_present_bogbean2019 <- bogbean2019[, colSums(bogbean2019 != 0) > 0]

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ncol = # species present
ncol(sp_present_bogbean1979) #30
ncol(sp_present_bogbean1999) #36
ncol(sp_present_bogbean2019) #34

##~~~~~~~~~~~~~~~~~~~
# Alpha diversity
##~~~~~~~~~~~~~~~~~~~

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
sp_plot_bogbean1979 <- apply(sp_present_bogbean1979 >0,1,sum)
#[1]  12 16 12 11  9 17 12 16 18 19 15 15  6  7 11 14 10 10 14

sp_plot_bogbean1999 <- apply(sp_present_bogbean1999 >0,1,sum)
#[1]  12 14 13 13 14 17 13 10 12 10 12 12 12 10  6  6  7 14

sp_plot_bogbean2019 <- apply(sp_present_bogbean2019 >0,1,sum)
#[1]  10 11  8 11 12 11 11 10 10  7 12 11  9 14 12  9 10  7 11 14 11 12  9 11  6 11 12 11

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# species density = mean # sp
mean(sp_plot_bogbean1979) #10.84
mean(sp_plot_bogbean1999) #11.50
mean(sp_plot_bogbean2019) #10.46

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#sd of species density
sd(sp_plot_bogbean1979) #3.61
sd(sp_plot_bogbean1999) #2.92
sd(sp_plot_bogbean2019) #1.90

##~~~~~~~~~~~~~~~~~~~
# Beta diversity: total n species divided by density
##~~~~~~~~~~~~~~~~~~~

ncol(sp_present_bogbean1979)/mean(sp_plot_bogbean1979) #2.77
ncol(sp_present_bogbean1999)/mean(sp_plot_bogbean1999) #3.13
ncol(sp_present_bogbean2019)/mean(sp_plot_bogbean2019) #3.25
```

RQ3: "What is the total community species turnover, and which species gained or lost are driving changes within each assemblage’s diversity?"
```{r turnover}
# https://rdrr.io/cran/codyn/man/turnover.html
#Note: turnover requires a column for abundance (not species presence/absence), here, Cover_class. "Cover_class" is a coarse abundance value where 0 = absent, 1 = 1-25% cover, 2 = 25-50% cover, etc. Because I want to understand turnover within the assemblage, I average all cover for a given species within that assemblage (across all quadrats (plots))

## Note: all calculations are comparison of 1979-1999, and 1999-2019. 


 # Calculate relative total turnover within replicates
total.res <- turnover(df = join_40, 
                     time.var = "Year", 
                     species.var = "Species", 
                     abundance.var = "mean_cover", 
                     replicate.var = "Assemblage", 
                     metric = "total")
kable(total.res)
 
# Calculate relative species appearances within replicates
appear.res <- turnover(df=join_40, 
               time.var = "Year", 
               species.var = "Species",
               abundance.var = "mean_cover",  
               replicate.var="Assemblage", 
               metric="appearance")

# Calculate relative species disappearances within replicates
 disappear.res <- turnover(df=join_40, 
               time.var = "Year", 
               species.var = "Species",
               abundance.var = "mean_cover",  
               replicate.var="Assemblage",
               metric="disappearance")

 turnover <- total.res %>% 
            left_join(appear.res, by = c("Assemblage", "Year")) %>% 
            left_join(disappear.res, by = c("Assemblage", "Year")) 
 
 turnover <- turnover[, c("Assemblage", "Year", "appearance", "disappearance", "total")]
 
# kable(turnover, 
#    col.names = c("Assemblage", "Year", "Species \nAppearance", "Species \nDisappearance", "Total \nTurnover"),
#    align = "c",  digits = 2) %>% #booktabs = TRUE,
#    #row_spec(1:6, color = 'black') %>%
#    kable_styling(font_size = 18, latex_options = "scale_down") %>%
#     save_kable("turnover.jpeg")
 

 
```

```{r indicator bar charts}


# Menyanthes_trifoliata, Mentha_aquatica, Lythrum_salicaria/Sidalcea_hendersonii, Phalaris_arundinacea/Iris_pseudacorus
freq.bogbean <- join_40 %>% 
   filter(Assemblage == "Bogbean") %>% 
   filter(Species %in% c("Menyanthes_trifoliata", "Mentha_aquatica", "Sidalcea_hendersonii", "Lythrum_salicaria", "Phalaris_arundinacea", "Iris_pseudacorus")) %>% 
   mutate(Species = factor(Species, 
                           levels = c("Menyanthes_trifoliata", "Mentha_aquatica", "Sidalcea_hendersonii", "Lythrum_salicaria", "Phalaris_arundinacea", "Iris_pseudacorus"), 
                           labels = c("Menyanthes_trifoliata", "Mentha_aquatica", "Sidalcea_hendersonii", "Lythrum_salicaria", "Phalaris_arundinacea", "Iris_pseudacorus"))) %>% 
   mutate(Species = recode(Species, "Menyanthes_trifoliata" = "Menyanthes trifoliata", "Mentha_aquatica" = "Mentha aquatica*", "Lythrum_salicaria" = "Lythrum salicaria*", "Sidalcea_hendersonii" = "Sidalcia hendersonii", "Phalaris_arundinacea" = "Phalaris arundinaceae*", "Iris_pseudacorus" = "Iris pseudacorus*"))

# FEAR, Salix_lasiandra (tip for Disc), LYSA/SIHE, PHAR/Iris_pseudacorus 
freq.fescue <- join_40 %>% 
   filter(Assemblage == "Fescue") %>% 
   filter(Species %in% c("Festuca_arundinacea", "Salix_lasiandra", "Sidalcea_hendersonii", "Lythrum_salicaria", "Phalaris_arundinacea", "Iris_pseudacorus")) %>% 
   mutate(Species = factor(Species, 
                           levels = c("Festuca_arundinacea", "Salix_lasiandra", "Sidalcea_hendersonii", "Lythrum_salicaria", "Phalaris_arundinacea", "Iris_pseudacorus"), 
                           labels = c("Festuca_arundinacea", "Salix_lasiandra", "Sidalcea_hendersonii", "Lythrum_salicaria", "Phalaris_arundinacea", "Iris_pseudacorus"))) %>% 
  mutate(Species = recode(Species, "Festuca_arundinacea" = "Festuca arundinacea*", "Salix_lasiandra" = "Salix lucida",  "Sidalcea_hendersonii" = "Sidalcia hendersonii", "Lythrum_salicaria" = "Lythrum salicaria*", "Phalaris_arundinacea" = "Phalaris arundinaceae*", "Iris_pseudacorus" = "Iris pseudacorus*")) 

#CALY, Salix_lasiandra, LYTHE/SIHE, PHAR/Iris_pseudacorus
freq.sedge <- join_40 %>% 
   filter(Assemblage == "Sedge") %>% 
   filter(Species %in% c("Carex_lyngbyei", "Salix_lasiandra", "Sidalcea_hendersonii", "Lythrum_salicaria", "Phalaris_arundinacea",  "Iris_pseudacorus")) %>% 
   mutate(Species = factor(Species, 
                           levels = c("Carex_lyngbyei", "Salix_lasiandra", "Sidalcea_hendersonii", "Lythrum_salicaria", "Phalaris_arundinacea",  "Iris_pseudacorus"), 
                           labels = c("Carex_lyngbyei", "Salix_lasiandra", "Sidalcea_hendersonii", "Lythrum_salicaria", "Phalaris_arundinacea",  "Iris_pseudacorus"))) %>% 
   mutate(Species = recode(Species, "Carex_lyngbyei" = "Carex lyngbyei", "Salix_lasiandra" = "Salix lucida",  "Sidalcea_hendersonii" = "Sidalcia hendersonii", "Lythrum_salicaria" = "Lythrum salicaria*", "Phalaris_arundinacea" = "Phalaris arundinaceae*", "Iris_pseudacorus" = "Iris pseudacorus*"))



# brewer.pal(n = 9, name = "BuGn")
# "#F7FCFD" "#E5F5F9" "#CCECE6" "#99D8C9" "#66C2A4" "#41AE76" "#238B45" "#006D2C" "#00441B"
freq.col <- c("#00441B", "#238B45" ,"#66C2A4")

freq.bogbean$Year <- as.factor(freq.bogbean$Year)
col.bogbean <- ggplot(freq.bogbean, 
                      aes(x = Species, 
                          y = mean_cover, 
                          fill = Year)) +
  geom_bar(stat="identity", 
           position = position_dodge(width = 0.5)) + 
  theme_classic() +
  expand_limits(y = c(0,4)) +
  scale_y_continuous(breaks = seq(0,4,0.5)) +
  labs(y = "", x = "") +
  theme(axis.text.x = element_text(size = 12, angle = 30, hjust = 1), 
        strip.text = element_text(size = 12)) +
  scale_fill_manual(values = freq.col) +
  facet_grid(.~ Assemblage)
col.bogbean

freq.sedge$Year <- as.factor(freq.sedge$Year)
col.sedge <- ggplot(freq.sedge, 
                    aes(x = Species, 
                        y = mean_cover, 
                        fill = Year)) +
  geom_bar(stat="identity", 
           position = position_dodge(width = 0.5)) + 
  theme_classic() +
  expand_limits(y = c(0,4)) +
  scale_y_continuous(breaks = seq(0,4,0.5)) +
  labs(y = "", x = "") +
  theme(axis.text.x = element_text(size = 12, angle = 30, hjust = 1), 
        strip.text = element_text(size = 12)) +
  scale_fill_manual(values = freq.col) +
  facet_grid(.~ Assemblage)
col.sedge

freq.fescue$Year <- as.factor(freq.fescue$Year)
col.fescue <- ggplot(freq.fescue, 
                     aes(x = Species, 
                         y = mean_cover, 
                         fill = Year)) +
  geom_bar(stat="identity", 
           position = position_dodge(width = 0.5)) + 
  theme_classic() +
  expand_limits(y = c(0,4)) +
  scale_y_continuous(breaks = seq(0,4,0.5)) +
  labs(y = "", x = "") +
  theme(axis.text.x = element_text(size = 12, angle = 30, hjust = 1), 
        strip.text = element_text(size = 12)) +
  scale_fill_manual(values = freq.col) +
  facet_grid(.~ Assemblage)
col.fescue

freq.panel <- ggarrange(col.sedge, 
                        col.fescue, 
                        col.bogbean,
                        common.legend = TRUE, 
                        legend = "right",
                        ncol = 1, nrow = 3, align = "hv") 
freq.panel

freq.bar <- annotate_figure(freq.panel, 
                            left = text_grob("Cover abundance class (mean)", 
                                             size = 20, rot = 90, hjust = .5, vjust = 1), 
                            bottom = text_grob("Species of interest", 
                                               size = 20))
freq.bar

#ggsave("fig-frequency-bar.jpg", freq.bar, width = 12, height = 16)



```


```{r all sp mean cover}

year.col <- c("#00441B", "#238B45" ,"#66C2A4")

join_40$Year <- as.factor(join_40$Year)


mean.cover <- ggplot(join_40, aes(x = Species, y = mean_cover, fill = Year)) +
  geom_bar(stat="identity", position = position_dodge(width = 0.5)) + theme_classic() +
  labs(y = "", x = "") +
  theme(axis.text.x = element_text(size = 11, angle = 30, hjust = 1), strip.text = element_text(
    size = 12)) +
  scale_fill_manual(values = year.col) +
  facet_grid(Assemblage ~ .)
mean.cover


```


Additional analyses

Supplemental: effect of dropping plots on spatial autocorrelation to diversity metircs

Run with a minimal number of plots (10)

```{r bootstrap 10 plots - 2019}
#start with diversity groupings as determined by each year's cluster analysis & indicator species analysis: d2019_grps, d1999_grps, d1979_grps
# Sub sample 10 trials using 10 plots from each sub assemblage. 

grps.19 <-  d2019_grps %>% 
  mutate(Assemblage = ifelse(Association_Type2 == 1, "sedge",
                      ifelse(Association_Type2 == 2, "fescue", "bogbean"))) %>% 
  relocate(Assemblage, .after = Year) %>% 
  select(-Transect, -Association_Type2) 

# select 10 random plots from each assemblage type & calculate diversity components

##~~~~~~~~~~~
# Sedge
##~~~~~~~~~~~
sedge.19 <- grps.19 %>% 
  filter(Assemblage == "sedge") %>% 
  select(-Assemblage, -Year)

#select 10 random plots
rando.sedge19 <- sedge.19[sample(nrow(sedge.19), 10), ]
unique(rando.sedge19$Quadrat)

# Trial 1 = c(57, 60, 45,  3, 50, 18, 53, 56, 55, 64)
# Trial 2 = c(13, 10,  8, 15, 52,  2, 70,  3,  9, 41)
# Trial 3 = c(60, 41, 55, 52, 50, 40, 57, 51, 70, 64)
# Trial 4 = c(40, 10, 58, 61, 51,  2, 56,  8, 41, 16)
# Trial 5 = c(16, 56, 54,  3, 18,  9, 53, 55, 52, 14)
# Trial 6 = c(70, 50, 19, 55, 18, 16, 56, 52, 61, 14)
# Trial 7 = c(3, 51, 18, 74, 15, 19, 53,  2, 13,  9)
# Trial 8 = c(16,  8, 40, 54, 14,  3, 60, 61, 56, 19)
# Trial 9 = c(58, 57, 16, 56, 13, 53, 70, 52, 60,  9)
# Trial 10 = c(64, 14, 74, 45,  9, 2, 10, 40, 60, 58)


r.sedge19 <- rando.sedge19 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.sedge.19 <- r.sedge19[, colSums(r.sedge19 != 0) > 0]
ncol(sp.sedge.19) 
# Trial 1 = 21
# Trial 2 = 27
# Trial 3 = 26
# Trial 4 = 24
# Trial 5 = 27
# Trial 6 = 29
# Trial 7 = 23
# Trial 8 = 27
# Trial 9 = 25
# Trial 10 = 24


# alpha diversity
pa.sedge.19 <- apply(sp.sedge.19 >0,1,sum) # convert cover value to presence/absence
mean(pa.sedge.19) 
#Trial 1 = 7.2
# Trial 2 = 8.4
# Trial 3 = 7.8
# Trial 4 = 7.5
# Trial 5 = 7.9
# Trial 6 = 8.4
# Trial 7 = 7.8
# Trial 8 = 7.7
# Trial 9 = 6.9
# Trial 10 = 7.5



sd(pa.sedge.19) 
#Trial 1 = 2.90
# Trial 2 = 2.01
# Trial 3 = 2.35
# Trial 4 = 3.10
# Trial 5 = 3.25
# Trial 6 = 3.03
# Trial 7 = 2.25
# Trial 8 = 3.13
# Trial 9 = 2.69
# Trial 10 = 2.95


# beta diversity
ncol(sp.sedge.19)/mean(pa.sedge.19) 
# Trial 1 = 2.92
# Trial 2 = 3.21
# Trial 3 = 3.33
# Trial 4 = 3.2
# Trial 5 = 3.42
# Trial 6 = 3.45
# Trial 7 = 2.95
# Trial 8 = 3.51
# Trial 9 = 3.62
# Trial 10 = 3.2


##~~~~~~~~~~~
# Fescue
##~~~~~~~~~~~

fescue.19 <- grps.19 %>% 
  filter(Assemblage == "fescue") %>% 
  select(-Assemblage, -Year)

#select 10 random plots
rando.fescue19 <- fescue.19[sample(nrow(fescue.19), 10), ]
unique(rando.fescue19$Quadrat)

# Trial 1 = c(28, 68, 65, 66,  1, 29, 43, 31, 69, 44)
# Trial 2 = c(32, 68, 66, 17, 28, 31, 30, 65, 44,  1)
# Trial 3 = c(69,  1, 31, 68, 44, 12, 43, 67, 66, 17)
# Trial 4 = c(68, 12, 17, 29, 59,  1, 31, 67, 43, 69)
# Trial 5 = c(12, 32, 69, 17, 31, 65, 44, 30, 67, 46)
# Trial 6 = c(32, 44, 30, 67, 66, 43,  1, 47, 69, 29)
# Trial 7 = c(44, 28, 30, 31, 46, 17, 12, 59, 65,  1)
# Trial 8 = c(29, 32, 67, 59, 31, 46, 12, 44, 66, 47)
# Trial 9 = c(31, 30, 29, 32, 66, 43, 46, 12, 67, 28)
# Trial 10 = c(17, 68, 44,  1, 59, 31, 29, 12, 30, 65)


r.fescue19 <- rando.fescue19 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.fescue.19 <- r.fescue19[, colSums(r.fescue19 != 0) > 0]
ncol(sp.fescue.19) 
# Trial 1 = 17
# Trial 2 = 20
# Trial 3 = 22
# Trial 4 = 21
# Trial 5 = 23
# Trial 6 = 19
# Trial 7 = 23
# Trial 8 = 25
# Trial 9 = 24
# Trial 10 = 23



# alpha diversity
pa.fescue.19 <- apply(sp.fescue.19 >0,1,sum) # convert cover value to presence/absence
mean(pa.fescue.19) 
# Trial 1 = 4.7
# Trial 2 = 5.5
# Trial 3 = 5.4
# Trial 4 = 5.2
# Trial 5 = 7.0
# Trial 6 = 5.3
# Trial 7 = 5.9
# Trial 8 = 6.5
# Trial 9 = 6.7
# Trial 10 = 4.9



sd(pa.fescue.19) 
# Trial 1 = 2.00 
# Trial 2 = 2.17
# Trial 3 = 2.59
# Trial 4 = 2.78
# Trial 5 = 2.83
# Trial 6 = 2.71
# Trial 7 = 2.88
# Trial 8 = 3.31
# Trial 9 = 3.23
# Trial 10 = 2.42


# beta diversity
ncol(sp.fescue.19)/mean(pa.fescue.19) 
# Trial 1 = 3.62
# Trial 2 = 3.64
# Trial 3 = 4.07
# Trial 4 = 4.04
# Trial 5 = 3.29
# Trial 6 = 3.58
# Trial 7 = 3.90
# Trial 8 = 3.85
# Trial 9 = 3.58
# Trial 10 = 4.69


##~~~~~~~~~~~
# Bogbean
##~~~~~~~~~~~


bogbean.19 <- grps.19 %>% 
  filter(Assemblage == "bogbean") %>% 
  select(-Assemblage, -Year)

#select 10 random plots
rando.bogbean19 <- bogbean.19[sample(nrow(bogbean.19), 10), ]
unique(rando.bogbean19$Quadrat)

# Trial 1 = c(39,  6, 38,  4, 72, 25, 20, 35, 11,  7)
# Trial 2 = c(72,  4, 38, 48, 49, 21, 20, 71, 35, 23)
# Trial 3 = c(11, 37, 71, 42, 27, 24, 20, 21, 35, 63)
# Trial 4 = c(49, 23, 71, 38, 34,  5, 26, 22,  7, 42)
# Trial 5 = c(71, 24, 39, 21, 42, 26, 23,  5, 49, 20)
# Trial 6 = c(23, 49, 42,  4, 48,  6, 63, 72, 71, 39)
# Trial 7 = c(33, 26, 11, 49,  6, 23, 73, 42, 71, 63)
# Trial 8 = c(25,  5, 20,  4, 39, 62, 34, 23, 63, 49)
# Trial 9 = c(34, 36, 38, 35, 26, 71, 37,  5, 23, 39)
# Trial 10 = c(22, 24,  5, 63, 33, 11, 37, 21, 48, 34)



r.bogbean19 <- rando.bogbean19 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.bogbean.19 <- r.bogbean19[, colSums(r.bogbean19 != 0) > 0]
ncol(sp.bogbean.19) 
# Trial 1 = 24 
# Trial 2 = 23
# Trial 3 = 22
# Trial 4 = 27
# Trial 5 = 24
# Trial 6 = 27
# Trial 7 = 27
# Trial 8 = 28
# Trial 9 = 26
# Trial 10 = 29



# alpha diversity
pa.bogbean.19 <- apply(sp.bogbean.19 >0,1,sum) # convert cover value to presence/absence
mean(pa.bogbean.19) 
#Trial 1 = 11.00 
# Trial 2 = 10.6
# Trial 3 = 9.4
# Trial 4 = 10.7
# Trial 5 = 10.6
# Trial 6 = 10.3
# Trial 7 = 10.3
# Trial 8 = 10.6
# Trial 9 = 10.6
# Trial 10 = 10.2


sd(pa.bogbean.19) 
#Trial 1 = 1.70 
# Trial 2 = 1.07
# Trial 3 = 2.12
# Trial 4 = 0.82
# Trial 5 = 1.78
# Trial 6 = 2.26
# Trial 7 = 2.21
# Trial 8 = 2.12
# Trial 9 = 1.84
# Trial 10 = 2.66


# beta diversity
ncol(sp.bogbean.19)/mean(pa.bogbean.19) 
# Trial 1 = 2.18
# Trial 2 = 2.17
# Trial 3 = 2.34
# Trial 4 = 2.52
# Trial 5 = 2.26
# Trial 6 = 2.62
# Trial 7 = 2.62
# Trial 8 = 2.64
# Trial 9 = 2.45
# Trial 10 = 2.84


##~~~~~~~~~~~
# Total
##~~~~~~~~~~~

rando.19 <- lst(rando.sedge19, rando.fescue19, rando.bogbean19) %>% 
  reduce(full_join) %>% 
  replace(is.na(.), 0) %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.rand.19 <- rando.19[, colSums(rando.19 != 0) > 0]
ncol(sp.rand.19) 
# Trial 1 = 35 
# Trial 2 = 36
# Trial 3 = 37
# Trial 4 = 36
# Trial 5 = 37
# Trial 6 = 38
# Trial 7 = 38
# Trial 8 = 40
# Trial 9 = 36
# Trial 10 = 40



# alpha diversity
pa.19 <- apply(sp.rand.19 >0,1,sum) # convert cover value to presence/absence
mean(pa.19) 
#Trial 1 = 7.63 
# Trial 2 = 8.12
# Trial 3 = 7.53
# Trial 4 = 7.8
# Trial 5 = 8.5
# Trial 6 = 8.0
# Trial 7 = 8.0
# Trial 8 = 8.27
# Trial 9 = 8.07
# Trial 10 = 7.53



sd(pa.19) 
#Trial 1 = 3.42 
# Trial 2 = 2.76
# Trial 3 = 2.83
# Trial 4 = 3.29
# Trial 5 = 3.03
# Trial 6 = 3.33
# Trial 7 = 3.01
# Trial 8 = 3.30
# Trial 9 = 3.14
# Trial 10 = 3.40


# beta diversity
ncol(sp.rand.19)/mean(pa.19) 
#Trial 1 = 4.58 
# Trial 2 = 4.41
# Trial 3 = 4.91
# Trial 4 = 4.61
# Trial 5 = 4.35
# Trial 6 = 4.75
# Trial 7 = 4.75
# Trial 8 = 4.84
# Trial 9 = 4.46
# Trial 10 = 5.31


```

```{r bootstrap 10 plots - 1999}
#start with diversity groupings as determined by each year's cluster analysis & indicator species analysis: d2019_grps, d1999_grps, d1979_grps
# Sub sample 10 trials using 10 plots from each sub assemblage. 

grps.99 <-  d1999_grps %>%
  mutate(Assemblage = ifelse(Association_Type2 == 1, "sedge",
                      ifelse(Association_Type2 == 2, "fescue", "bogbean"))) %>%
  relocate(Assemblage, .after = Year) %>%
  select(-Transect, -Association_Type2)


# select 10 random plots from each assemblage type & calculate diversity components

##~~~~~~~~~~~
# Sedge
##~~~~~~~~~~~
sedge.99 <- grps.99 %>% 
  filter(Assemblage == "sedge") %>% 
  select(-Assemblage, -Year)

#select 10 random plots
rando.sedge99 <- sedge.99[sample(nrow(sedge.99), 10), ]
unique(rando.sedge99$Quadrat)
# Trial 1 = c(85,  73,  29,  39,  42,  31,  87,  23,  28, 100)
# Trial 2 = c(24,  27,  41,  88,  87,  42,  39,  94, 101, 100)
# Trial 3 = c(31, 12, 41, 30, 96, 72, 26, 69, 73, 44)
# Trial 4 = c(32,  72,  29,  40, 101,  46, 31,  24,  43,  88)
# Trial 5 = c(24, 28, 85, 46, 43, 29, 32, 73, 30, 95)
# Trial 6 = c(42,  44,  69,  40,  31,  43,  72,  27,  26, 108)
# Trial 7 = c(40,  12, 100,  87,  23,  73,  43,  95,  88,  32)
# Trial 8 = c(96,  31, 108, 24, 100,  85,  22 , 46,  40,  32)
# Trial 9 = c(96, 39, 30, 22, 27, 69, 40, 44, 73, 42)
# Trial 10 = c(31,  27,  73,  22, 108,  88,  87,  39,  28,  32)


r.sedge99 <- rando.sedge99 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.sedge.99 <- r.sedge99[, colSums(r.sedge99 != 0) > 0]
ncol(sp.sedge.99) 
# Trial 1 =  29
# Trial 2 = 21
# Trial 3 = 29
# Trial 4 = 29
# Trial 5 = 29
# Trial 6 = 27
# Trial 7 = 29
# Trial 8 = 30
# Trial 9 = 29
# Trial 10 = 26



# alpha diversity
pa.sedge.99 <- apply(sp.sedge.99 >0,1,sum) # convert cover value to presence/absence
mean(pa.sedge.99) 
# Trial 1 = 8.8
# Trial 2 = 7.7
# Trial 3 = 8.7
# Trial 4 = 7.9
# Trial 5 = 8.1
# Trial 6 = 8.4
# Trial 7 = 7.9
# Trial 8 = 8.3
# Trial 9 = 8.6
# Trial 10 = 7.2



sd(pa.sedge.99) 
#Trial 1 = 1.93
# Trial 2 = 1.70
# Trial 3 = 2.00
# Trial 4 = 1.66
# Trial 5 = 2.18
# Trial 6 = 2.46
# Trial 7 = 1.20
# Trial 8 = 2.31
# Trial 9 = 2.07
# Trial 10 = 1.14



# beta diversity
ncol(sp.sedge.99)/mean(pa.sedge.99) 
#Trial 1 = 3.30
# Trial 2 = 2.73
# Trial 3 = 3.33
# Trial 4 = 3.67
# Trial 5 = 3.58
# Trial 6 = 3.21
# Trial 7 = 3.67
# Trial 8 = 3.61
# Trial 9 = 3.37
# Trial 10 = 3.61


##~~~~~~~~~~~
# Fescue
##~~~~~~~~~~~

fescue.99 <- grps.99 %>% 
  filter(Assemblage == "fescue") %>% 
  select(-Assemblage, -Year)

#select 10 random plots
rando.fescue99 <- fescue.99[sample(nrow(fescue.99), 10), ]
unique(rando.fescue99$Quadrat)
#Trial 1 = c(25,  21,  56,  66,  47,  57,  48,  38,  10, 106)
# Trial 2 = c(50, 57, 10, 74, 75, 56, 59, 66, 86, 21)
# Trial 3 = c(50, 107,  75, 106,  10,  62,  66,  21,  93,  60)
# Trial 4 = c(48, 56, 10, 58, 57, 63,  9, 60, 75, 67)
# Trial 5 = c(65,  49,  10,  56,  75, 107, 106,  76,  20, 37)
# Trial 6 = c(61, 56, 66, 11, 55, 57, 68, 58, 63, 50)
# Trial 7 = c(11, 64, 76, 55, 38, 63, 62, 75, 58, 59)
# Trial 8 = c(55, 20, 59, 25, 57, 60,  9, 63, 93, 86)
# Trial 9 = c(86, 58, 10, 75, 20, 93, 55, 21, 60, 68)
# Trial 10 = c(67, 58, 48, 11,  9, 55, 47, 66, 57, 76)



r.fescue99 <- rando.fescue99 %>% 
  select(-Quadrat)


# total number of species = ncol > 0
sp.fescue.99 <- r.fescue99[, colSums(r.fescue99 != 0) > 0]
ncol(sp.fescue.99) 
#Trial 1 = 31
# Trial 2 = 32
# Trial 3 = 31
# Trial 4 = 30
# Trial 5 = 31
# Trial 6 = 32
# Trial 7 = 30
# Trial 8 = 27
# Trial 9 = 28
# Trial 10 = 32


# alpha diversity
pa.fescue.99 <- apply(sp.fescue.99 >0,1,sum) # convert cover value to presence/absence
mean(pa.fescue.99) 
#Trial 1 = 10.3
# Trial 2 = 9.1
# Trial 3 = 8.00
# Trial 4 = 10.7
# Trial 5 = 8.2
# Trial 6 = 12.9
# Trial 7 = 9.1
# Trial 8 = 8.3
# Trial 9 = 8.6
# Trial 10 = 11.2


sd(pa.fescue.99) 
#Trial 1 = 4.95
# Trial 2 = 3.67
# Trial 3 = 2.94
# Trial 4 = 4.45
# Trial 5 = 3.88
# Trial 6 = 2.82
# Trial 7 = 3.21
# Trial 8 = 3.06
# Trial 9 = 3.24
# Trial 10 = 3.91


# beta diversity
ncol(sp.fescue.99)/mean(pa.fescue.99) 
#Trial 1 = 3.01
# Trial 2 = 3.52
# Trial 3 = 3.88
# Trial 4 = 2.80
# Trial 5 = 3.78
# Trial 6 = 2.48
# Trial 7 = 3.30
# Trial 8 = 3.25
# Trial 9 = 3.26
# Trial 10 = 2.86


##~~~~~~~~~~~
# Bogbean
##~~~~~~~~~~~

bogbean.99 <- grps.99 %>% 
  filter(Assemblage == "bogbean") %>% 
  select(-Assemblage, -Year)

#select 10 random plots
rando.bogbean99 <- bogbean.99[sample(nrow(bogbean.99), 10), ]
unique(rando.bogbean99$Quadrat)
#Trial 1 = c( 83,  13,  16, 15, 104,  14, 102,  78,  80, 103)
# Trial 2 = c(81,  82,  98,  15, 103,  83,  77,  16, 102,  84)
# Trial 3 = c(104,  84,  77, 103,  78, 105,  14,  80,  15,  99)
# Trial 4 = c(15, 103,  82,  84, 104,  78,  14,  13, 80,  77)
# Trial 5 = c(102, 103,  79,  84,  83,  98,  13,  77, 105,  78)
# Trial 6 = c(15,  77,  79, 102,  82,  98,  14,  81, 105,  80)
# Trial 7 = c(83,  79,  77,  81, 104, 102,  78,  80, 105,  16)
# Trial 8 = c(103,  99,  98, 104,  80,  78,  83,  15,  77, 102)
# Trial 9 = c( 103,  13,  82,  99,  77,  15, 102,  79, 105,  98)
# Trial 10 = c(13,  79, 103, 104, 105,  99,  16,  81,  82,  83)



r.bogbean99 <- rando.bogbean99 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.bogbean.99 <- r.bogbean99[, colSums(r.bogbean99 != 0) > 0]
ncol(sp.bogbean.99) 
#Trial 1 = 31
# Trial 2 = 31
# Trial 3 = 32
# Trial 4 = 32
# Trial 5 = 32
# Trial 6 = 33
# Trial 7 = 33
# Trial 8 = 31
# Trial 9 = 34
# Trial 10 = 34


# alpha diversity
pa.bogbean.99 <- apply(sp.bogbean.99 >0,1,sum) # convert cover value to presence/absence
mean(pa.bogbean.99) 
#Trial 1 = 11
# Trial 2 = 11
# Trial 3 = 11.7
# Trial 4 = 11.5
# Trial 5 = 11.8
# Trial 6 = 11.8
# Trial 7 = 11.8
# Trial 8 = 10.7
# Trial 9 = 11
# Trial 10 = 10.9

sd(pa.bogbean.99) 
#Trial 1 = 3.68
# Trial 2 = 2.83
# Trial 3 = 3.43
# Trial 4 = 3.34
# Trial 5 = 3.43
# Trial 6 = 2.53
# Trial 7 = 3.33
# Trial 8 = 3.62
# Trial 9 = 2.98
# Trial 10 = 2.64


# beta diversity
ncol(sp.bogbean.99)/mean(pa.bogbean.99) 
#Trial 1 = 2.82
# Trial 2 = 2.82
# Trial 3 = 2.74
# Trial 4 = 2.78
# Trial 5 = 2.71
# Trial 6 = 2.80
# Trial 7 = 2.80
# Trial 8 = 2.90
# Trial 9 = 3.09
# Trial 10 = 3.12


##~~~~~~~~~~~
# Total
##~~~~~~~~~~~

rando.99 <- lst(rando.sedge99, rando.fescue99, rando.bogbean99) %>% 
  reduce(full_join) %>% 
  replace(is.na(.), 0) %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.rand.99 <- rando.99[, colSums(rando.99 != 0) > 0]
ncol(sp.rand.99) 
#Trial 1 = 39
# Trial 2 = 38
# Trial 3 = 41
# Trial 4 = 42
# Trial 5 = 39
# Trial 6 = 41
# Trial 7 = 42
# Trial 8 = 39
# Trial 9 = 40
# Trial 10 = 40


# alpha diversity
pa.99 <- apply(sp.rand.99 >0,1,sum) # convert cover value to presence/absence
mean(pa.99) 
#Trial 1 = 10.03
# Trial 2 = 9.27
# Trial 3 = 9.47
# Trial 4 = 10.03
# Trial 5 = 9.37
# Trial 6 = 11.03
# Trial 7 = 9.6
# Trial 8 = 9.1
# Trial 9 = 9.4
# Trial 10 = 9.77


sd(pa.99)
#Trial 1 = 3.72
# Trial 2 = 3.07
# Trial 3 = 3.20
# Trial 4 = 3.60
# Trial 5 = 3.59
# Trial 6 = 3.05
# Trial 7 = 3.14
# Trial 8 = 3.16
# Trial 9 = 2.94
# Trial 10 = 3.28


# beta diversity
ncol(sp.rand.99)/mean(pa.99) 
#Trial 1 = 3.89
# Trial 2 = 4.10
# Trial 3 = 4.33
# Trial 4 = 4.19
# Trial 5 = 4.16
# Trial 6 = 3.72
# Trial 7 = 4.38
# Trial 8 = 4.29
# Trial 9 = 4.26
# Trial 10 = 4.10
```

```{r bootstrap 10 plots - 1979 }
#start with diversity groupings as determined by each year's cluster analysis & indicator species analysis: d2019_grps, d1999_grps, d1979_grps
# Sub sample 10 trials using 10 plots from each sub assemblage. 


#Filter each year; n quadrat = nrow
grps.79 <-  d1979_grps %>%
  mutate(Assemblage = ifelse(Association_Type2 == 1, "sedge",
                      ifelse(Association_Type2 == 2, "fescue", "bogbean"))) %>%
  relocate(Assemblage, .after = Year) %>%
  select(-Transect, -Association_Type, -Association_Type2, -Association_Subtype)

sedge.79 <- grps.79 %>% 
  filter(Assemblage == "sedge") %>% 
  select(-Assemblage, -Year)



#select 10 random plots
rando.sedge79 <- sedge.79[sample(nrow(sedge.79), 10), ]
unique(rando.sedge79$Quadrat)
# Trial 1 = c(26, 88, 47, 29, 59, 57, 60, 72, 22, 31)
# Trial 2 = c(25,  47,  48,  20,  57,  59,  72,  58, 108,  86)
# Trial 3 = c(73, 58, 59, 55, 57, 48, 30, 27, 87, 88)
# Trial 4 = c(96, 97, 29, 22, 86, 93, 57, 56, 49, 47)
# Trial 5 = c(32, 49, 26, 93, 31, 59, 69, 48, 47, 88)
# Trial 6 = c(31,  27,  73,  22, 108,  88,  87,  39,  28,  32)
# Trial 7 = c(24,  60,  26, 108,  94,  28,  88,  97,  23,  31)
# Trial 8 = c(95, 108,  25,  87,  22,  58,  49,  55,  27,  48)
# Trial 9 = c(86, 88, 20, 26, 72, 48, 93, 60, 69, 94)
# Trial 10 = c(47, 56, 87, 32, 25, 73, 58, 28, 93, 27)




r.sedge79 <- rando.sedge79 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.sedge.79 <- r.sedge79[, colSums(r.sedge79 != 0) > 0]
ncol(sp.sedge.79) 
# Trial 1 = 27
# Trial 2 = 29
# Trial 3 = 24
# Trial 4 = 29
# Trial 5 = 26
# Trial 6 = 26
# Trial 7 = 27
# Trial 8 = 26
# Trial 9 = 33
# Trial 10 = 27


# alpha diversity
pa.sedge.79 <- apply(sp.sedge.79 >0,1,sum) # convert cover value to presence/absence
mean(pa.sedge.79) 
# Trial 1 = 8.7
# Trial 2 = 8.6
# Trial 3 = 8.8
# Trial 4 = 9.5
# Trial 5 = 8.9
# Trial 6 = 7.2
# Trial 7 = 8.5
# Trial 8 = 8.4
# Trial 9 = 9.6
# Trial 10 = 8.9


sd(pa.sedge.79) 
# Trial 1 = 2.67
# Trial 2 = 2.37
# Trial 3 = 1.69
# Trial 4 = 2.46
# Trial 5 = 2.81
# Trial 6 = 1.14
# Trial 7 = 3.10
# Trial 8 = 2.72
# Trial 9 = 2.72
# Trial 10 = 2.23


# beta diversity
ncol(sp.sedge.79)/mean(pa.sedge.79)
# Trial 1 = 3.10
# Trial 2 = 3.37
# Trial 3 = 2.73
# Trial 4 = 3.05
# Trial 5 = 2.92
# Trial 6 = 3.61
# Trial 7 = 3.18
# Trial 8 = 3.10
# Trial 9 = 3.44
# Trial 10 = 3.03


##~~~~~~~~~~~
# Fescue
##~~~~~~~~~~~

fescue.79 <- grps.79 %>% 
  filter(Assemblage == "fescue") %>% 
  select(-Assemblage, -Year)

#select 10 random plots
rando.fescue79 <- fescue.79[sample(nrow(fescue.79), 10), ]
unique(rando.fescue79$Quadrat)
# Trial 1 = c(38,  40, 102,  46,  14,  77,  12,  75,  13,  16)
# Trial 2 = c(44,  16,  10, 101,  41,   9,  75,  43,  37,  77)
# Trial 3 = c(40,  43,   9, 102,  76, 104,  42,  41,  44, 103)
# Trial 4 = c(9,  77,  10,  40,  46,  39, 107,  44,  41,  11)
# Trial 5 = c(13,  43,  46,  15,  14,  21,  44, 106,  16, 101)
# Trial 6 = c(67, 58, 48, 11,  9, 55, 47, 66, 57, 76)
# Trial 7 = c(21, 44, 39, 12, 75, 40, 74, 46,  9, 15)
# Trial 8 = c(40, 107, 102,  75,  10,  12,  41,  76, 105,  74)
# Trial 9 = c(44,  10,  14,   9,  21,  43, 107,  11,  40,  46)
# Trial 10 = c(15, 44,  9, 38, 75, 10, 40, 39, 14, 41)




r.fescue79 <- rando.fescue79 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.fescue.79 <- r.fescue79[, colSums(r.fescue79 != 0) > 0]
ncol(sp.fescue.79) 
# Trial 1 = 37
# Trial 2 = 33
# Trial 3 = 39
# Trial 4 = 31
# Trial 5 = 37
# Trial 6 = 32
# Trial 7 = 35
# Trial 8 = 35
# Trial 9 = 37
# Trial 10 = 35


# alpha diversity
pa.fescue.79 <- apply(sp.fescue.79 >0,1,sum) # convert cover value to presence/absence
mean(pa.fescue.79) 
# Trial 1 = 11.5
# Trial 2 = 10.1
# Trial 3 = 12.8
# Trial 4 = 10.1
# Trial 5 = 11.7
# Trial 6 = 11.2
# Trial 7 = 11.0
# Trial 8 = 10.5
# Trial 9 = 10.1
# Trial 10 = 10.0



sd(pa.fescue.79) 
# Trial 1 = 3.78
# Trial 2 = 4.07
# Trial 3 = 4.49
# Trial 4 = 4.20
# Trial 5 = 2.50
# Trial 6 = 3.91
# Trial 7 = 3.94
# Trial 8 = 4.12
# Trial 9 = 3.93
# Trial 10 = 4.40


# beta diversity
ncol(sp.fescue.79)/mean(pa.fescue.79) 
# Trial 1 = 3.22
# Trial 2 = 3.27
# Trial 3 = 3.05
# Trial 4 = 3.07
# Trial 5 = 3.16
# Trial 6 = 2.86
# Trial 7 = 3.18
# Trial 8 = 3.33
# Trial 9 = 3.66
# Trial 10 = 3.5


##~~~~~~~~~~~
# Bogbean
##~~~~~~~~~~~


bogbean.79 <- grps.79 %>% 
  filter(Assemblage == "bogbean") %>% 
  select(-Assemblage, -Year)

#select 10 random plots
rando.bogbean79 <- bogbean.79[sample(nrow(bogbean.79), 10), ]
unique(rando.bogbean79$Quadrat)
# Trial 1 = c(100,  82,  79,  67,  78,  84,  63,  62,  64,  50)
# Trial 2 = c(67,  81,  62,  79,  83,  63,  85,  65,  66, 100)
# Trial 3 = c(79, 82, 83, 65, 50, 85, 67, 99, 66, 78)
# Trial 4 = c(61,  81,  79,  66,  99,  84,  80,  67,  78, 100)
# Trial 5 = c(79, 99, 63, 67, 82, 98, 81, 78, 65, 62)
# Trial 6 = c(13,  79, 103, 104, 105,  99,  16,  81,  82,  83)
# Trial 7 = c(67, 100,  84,  79,  83,  98,  62,  82,  66,  80)
# Trial 8 = c(98, 79, 65, 66, 81, 99, 67, 84, 62, 80)
# Trial 9 = c(85, 63, 78, 98, 84, 83, 61, 64, 62, 99)
# Trial 10 = c(64, 63, 78, 99, 81, 98, 67, 84, 61, 82)



r.bogbean79 <- rando.bogbean79 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.bogbean.79 <- r.bogbean79[, colSums(r.bogbean79 != 0) > 0]
ncol(sp.bogbean.79) 
# Trial 1 = 27
# Trial 2 = 29
# Trial 3 = 29
# Trial 4 = 27
# Trial 5 = 27
# Trial 6 = 34
# Trial 7 = 25
# Trial 8 = 25
# Trial 9 = 26
# Trial 10 = 24


# alpha diversity
pa.bogbean.79 <- apply(sp.bogbean.79 >0,1,sum) # convert cover value to presence/absence
mean(pa.bogbean.79) 
# Trial 1 = 10.8
# Trial 2 = 11.7
# Trial 3 = 11.1
# Trial 4 = 12.6
# Trial 5 = 11.4
# Trial 6 = 10.9
# Trial 7 = 10.2
# Trial 8 = 11.7
# Trial 9 = 9.8
# Trial 10 = 10.2



sd(pa.bogbean.79) 
# Trial 1 = 4.02
# Trial 2 = 3.40
# Trial 3 = 4.51
# Trial 4 = 2.91
# Trial 5 = 4.22
# Trial 6 = 2.64
# Trial 7 = 3.99
# Trial 8 = 3.13
# Trial 9 = 3.33
# Trial 10 = 3.82



# beta diversity
ncol(sp.bogbean.79)/mean(pa.bogbean.79) 
# Trial 1 = 2.50
# Trial 2 = 2.48
# Trial 3 = 2.61
# Trial 4 = 2.14
# Trial 5 = 2.37
# Trial 6 = 3.12
# Trial 7 = 2.45
# Trial 8 = 2.14
# Trial 9 = 2.65
# Trial 10 = 2.35


##~~~~~~~~~~~
# Total
##~~~~~~~~~~~

rando.79 <- lst(rando.sedge79, rando.fescue79, rando.bogbean79) %>% 
  reduce(full_join) %>% 
  replace(is.na(.), 0) %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.rand.79 <- rando.79[, colSums(rando.79 != 0) > 0]
ncol(sp.rand.79) 
# Trial 1 = 40
# Trial 2 = 42
# Trial 3 = 43
# Trial 4 = 40
# Trial 5 = 38
# Trial 6 = 40
# Trial 7 = 41
# Trial 8 = 42
# Trial 9 = 42
# Trial 10 = 40



# alpha diversity
pa.79 <- apply(sp.rand.79 >0,1,sum) # convert cover value to presence/absence
mean(pa.79) 
# Trial 1 = 10.33
# Trial 2 = 10.13
# Trial 3 = 10.9
# Trial 4 = 10.73
# Trial 5 = 10.67
# Trial 6 = 9.77
# Trial 7 = 9.9
# Trial 8 = 10.2
# Trial 9 = 9.83
# Trial 10 = 9.7


sd(pa.79) 
# Trial 1 = 3.62
# Trial 2 = 3.48
# Trial 3 = 4.03
# Trial 4 = 3.44
# Trial 5 = 3.40
# Trial 6 = 3.28
# Trial 7 = 3.73
# Trial 8 = 3.54
# Trial 9 = 3.25
# Trial 10 = 3.52


# beta diversity
ncol(sp.rand.79)/mean(pa.79) 
# Trial 1 = 3.87
# Trial 2 = 4.14
# Trial 3 = 3.94
# Trial 4 = 3.73
# Trial 5 = 3.56
# Trial 6 = 4.10
# Trial 7 = 4.14
# Trial 8 = 4.12
# Trial 9 = 4.27
# Trial 10 = 4.12


```

Bootstrap with minimum number of plots sampled per assemblage across datasets (18)
```{r bootstrap 18 plots - 2019}

grps.19 <-  d2019_grps %>% 
  mutate(Assemblage = ifelse(Association_Type2 == 1, "sedge",
                      ifelse(Association_Type2 == 2, "fescue", "bogbean"))) %>% 
  relocate(Assemblage, .after = Year) %>% 
  select(-Transect, -Association_Type2) 

# select 18 random plots from each assemblage type & calculate diversity components

##~~~~~~~~~~~
# Sedge
##~~~~~~~~~~~
sedge.19 <- grps.19 %>% 
  filter(Assemblage == "sedge") %>% 
  select(-Assemblage, -Year)

#select 10 random plots
rando18.sedge19 <- sedge.19[sample(nrow(sedge.19), 18), ]
unique(rando18.sedge19$Quadrat)

# Trial 1 = c(54 52  2 16 55 64 13 19 50 41 56 61 74  3 58 57  8 14)
# Trial 2 = c(18 45 15 53  9 74 13 55 56 14 61 58 41  3 52  8 64 60)
# Trial 3 = c(40 54 58 60 15 61 64 52  8 41 10 18  9 14 13  2 45 56)
# Trial 4 = c(14 55 57 60 53 41  9 19  3 50 74 15 58 54  2 70 64 18)
# Trial 5 = c(56  9 70 15  8 74 16 53 14 13 64 19 57 50 61 55 10  3)
# Trial 6 = c(9 53 61  3 10 55 64 40 45 56 18 70  2 52 54 15 57  8)
# Trial 7 = c(2 55 57 58 53  9 19  8 51 18 70 13 40 64 52 50 14 56)
# Trial 8 = c(60  8 74 10 53 40 18 56 70 61  2 64 41 19  3 15 57 54)
# Trial 9 = c(18 16 13  2 70 41 40 15 10 53 51 50 74 60 54 19 57  9)
# Trial 10 = c(13 52 50 55  8 61 53  9 18 14  2 19 54 41 57 10 40 45)



r18.sedge19 <- rando18.sedge19 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.sedge18.19 <- r18.sedge19[, colSums(r18.sedge19 != 0) > 0]
ncol(sp.sedge18.19) 
# Trial 1 = 34
# Trial 2 = 31
# Trial 3 = 29
# Trial 4 = 32
# Trial 5 = 31
# Trial 6 = 28
# Trial 7 = 33
# Trial 8 = 26
# Trial 9 = 30
# Trial 10 = 34


# alpha diversity
pa.sedge18.19 <- apply(sp.sedge18.19 >0,1,sum) # convert cover value to presence/absence
mean(pa.sedge18.19) 

# Trial 1 = 8.22
# Trial 2 = 7.89
# Trial 3 = 7.83
# Trial 4 = 8.0
# Trial 5 = 8.61
# Trial 6 = 8.06
# Trial 7 = 8.17
# Trial 8 = 8.06
# Trial 9 = 8.0
# Trial 10 = 9.0


sd(pa.sedge18.19) 
# Trial 1 = 2.96
# Trial 2 = 2.78
# Trial 3 = 2.43
# Trial 4 = 2.70
# Trial 5 = 2.75
# Trial 6 = 2.34
# Trial 7 = 2.41
# Trial 8 = 2.26
# Trial 9 = 2.43
# Trial 10 = 2.03


# beta diversity
ncol(sp.sedge18.19)/mean(pa.sedge18.19) 
# Trial 1 = 4.14
# Trial 2 = 3.93
# Trial 3 = 3.70
# Trial 4 = 4.0
# Trial 5 = 3.6
# Trial 6 = 3.48
# Trial 7 = 4.04
# Trial 8 = 3.23
# Trial 9 = 3.75
# Trial 10 = 3.78


##~~~~~~~~~~~
# Fescue
##~~~~~~~~~~~
## *** Values will always be the same bc there are exactly 18 plots; not running 10x. 
fescue.19 <- grps.19 %>% 
  filter(Assemblage == "fescue") %>% 
  select(-Assemblage, -Year)

#select 10 random plots
rando18.fescue19 <- fescue.19[sample(nrow(fescue.19), 18), ]
unique(rando18.fescue19$Quadrat)
# Trial 1 = c(65 69 31  1 47 32 46 59 44 30 17 12 66 29 67 43 68 28)
# Trial 2 = c(29 68 31  1 69 46 30 28 67 59 47 17 65 66 44 32 43 12)




r18.fescue19 <- rando18.fescue19 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.fescue18.19 <- r18.fescue19[, colSums(r18.fescue19 != 0) > 0]
ncol(sp.fescue18.19) 
# Trial 1 = 27
# Trial 2 = 27


# alpha diversity
pa.fescue18.19 <- apply(sp.fescue18.19 >0,1,sum) # convert cover value to presence/absence
mean(pa.fescue18.19) 
# Trial 1 = 5.83
# Trial 2 = 5.83


sd(pa.fescue18.19) 
# Trial 1 = 2.79
# Trial 2 = 2.79


# beta diversity
ncol(sp.fescue18.19)/mean(pa.fescue18.19) 
# Trial 1 = 4.63
# Trial 2 = 4.63

##~~~~~~~~~~~
# Bogbean
##~~~~~~~~~~~


bogbean.19 <- grps.19 %>% 
  filter(Assemblage == "bogbean") %>% 
  select(-Assemblage, -Year)

#select 10 random plots
rando18.bogbean19 <- bogbean.19[sample(nrow(bogbean.19), 18), ]
unique(rando18.bogbean19$Quadrat)
# Trial 1 = c(5 48  6  4 49 42 25 23 39 24 72 11 34 36 73 21 38 20)
# Trial 2 = c(62  6 33  7 71 25 21 22 23 72 49 27 11 34 42 36 63  4)
# Trial 3 = c(38 34 39 11 22  6 26 27 49 20 62 37 72 21 48 35 33 36)
# Trial 4 = c(39 72 49 25  4 63 62  5 21 26  6 11 34 73 35 24 33 27)
# Trial 5 = c(6 22 26 42 37 21 20 49  5 11 27 33 23 72 48 38  4 39)
# Trial 6 = c(5 48 63 36 21 25 27 22 24 20 71 73 38  6 11 33 42 49)
# Trial 7 = c(4 73 24 22 36 27 37 62 20 34 11  6 63 26 35 48 42 25)
# Trial 8 = c(26 20 42 63 62 34 21 22 24 48 33 11 39  7 23  6  5 71)
# Trial 9 = c(4 62 36  5 71 11 27 37 20 48 26 73 72 24 21 49 25 63)
# Trial 10 = c(21 22 34 62 33 36 63 11 42 27 39 35 48 25 72  6  4 38)


r18.bogbean19 <- rando18.bogbean19 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.bogbean18.19 <- r18.bogbean19[, colSums(r18.bogbean19 != 0) > 0]
ncol(sp.bogbean18.19) 
# Trial 1 = 31
# Trial 2 = 32
# Trial 3 = 32
# Trial 4 = 32
# Trial 5 = 32
# Trial 6 = 31
# Trial 7 = 31
# Trial 8 = 31
# Trial 9 = 29
# Trial 10 = 33

# alpha diversity
pa.bogbean18.19 <- apply(sp.bogbean18.19 >0,1,sum) # convert cover value to presence/absence
mean(pa.bogbean18.19) 
# Trial 1 = 10.78
# Trial 2 = 10.5
# Trial 3 = 10.72
# Trial 4 = 10.5
# Trial 5 = 10.72
# Trial 6 = 10.33
# Trial 7 = 9.94
# Trial 8 = 10.72
# Trial 9 = 10.17
# Trial 10 = 10.78


sd(pa.bogbean18.19) 
# Trial 1 = 1.63
# Trial 2 = 1.79
# Trial 3 = 1.87
# Trial 4 = 2.18
# Trial 5 = 1.81
# Trial 6 = 1.94
# Trial 7 = 1.89
# Trial 8 = 2.05
# Trial 9 = 1.86
# Trial 10 = 1.99


# beta diversity
ncol(sp.bogbean18.19)/mean(pa.bogbean18.19) 
# Trial 1 = 2.87
# Trial 2 = 3.05
# Trial 3 = 2.98
# Trial 4 = 3.05
# Trial 5 = 2.98
# Trial 6 = 3.0
# Trial 7 = 3.12
# Trial 8 = 2.89
# Trial 9 = 2.85
# Trial 10 = 3.06


##~~~~~~~~~~~
# Total
##~~~~~~~~~~~

rando18.19 <- lst(rando18.sedge19, rando18.fescue19, rando18.bogbean19) %>% 
  reduce(full_join) %>% 
  replace(is.na(.), 0) %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.rand18.19 <- rando18.19[, colSums(rando18.19 != 0) > 0]
ncol(sp.rand18.19) 

# Trial 1 = 43
# Trial 2 = 42
# Trial 3 = 41
# Trial 4 = 43
# Trial 5 = 41
# Trial 6 = 39
# Trial 7 = 43
# Trial 8 = 40
# Trial 9 = 42
# Trial 10 = 43



# alpha diversity
pa18.19 <- apply(sp.rand18.19 >0,1,sum) # convert cover value to presence/absence
mean(pa18.19) 

# Trial 1 = 8.28
# Trial 2 = 8.07
# Trial 3 = 8.13
# Trial 4 = 8.11
# Trial 5 = 8.39
# Trial 6 = 8.07
# Trial 7 = 7.98
# Trial 8 = 8.20
# Trial 9 = 8.0
# Trial 10 = 8.54


sd(pa18.19) 

# Trial 1 = 3.21
# Trial 2 = 3.12
# Trial 3 = 3.10
# Trial 4 = 3.17
# Trial 5 = 3.17
# Trial 6 = 2.98
# Trial 7 = 2.90
# Trial 8 = 3.09
# Trial 9 = 2.95
# Trial 10 = 3.06

# beta diversity
ncol(sp.rand18.19)/mean(pa18.19) 

# Trial 1 = 5.19
# Trial 2 = 5.20
# Trial 3 = 5.04
# Trial 4 = 5.30
# Trial 5 = 4.89
# Trial 6 = 4.83
# Trial 7 = 5.39
# Trial 8 = 4.88
# Trial 9 = 5.25
# Trial 10 = 5.04

```

```{r bootstrap 18 plots - 1999}

grps.99 <-  d1999_grps %>% 
  mutate(Assemblage = ifelse(Association_Type2 == 1, "sedge",
                      ifelse(Association_Type2 == 2, "fescue", "bogbean"))) %>% 
  relocate(Assemblage, .after = Year) %>% 
  select(-Transect, -Association_Type2) 

# select 18 random plots from each assemblage type & calculate diversity components

##~~~~~~~~~~~
# Sedge
##~~~~~~~~~~~
sedge.99 <- grps.99 %>% 
  filter(Assemblage == "sedge") %>% 
  select(-Assemblage, -Year)

#select 18 random plots
rando18.sedge99 <- sedge.99[sample(nrow(sedge.99), 18), ]
unique(rando18.sedge99$Quadrat)

# Trial 1 = c(108  23  26  24  95  22  87  94  29  30  39  42  12  73  72  28  43  88)
# Trial 2 = c( 88  26  44  97 101  27  24  23  30  29  85  43  41  95  31  87  12  69)
# Trial 3 = c(42  43  39  30  12  85  24  27  31  96  22  28 101  73 100  69  40  44)
# Trial 4 = c(29  12  39  73  85  28  24  31 101  88  97  22  46  32  41  23  26  27)
# Trial 5 = c(41 100  94  85  73  29 108  32  88  44  72  43  87  42  96  23  28  27)
# Trial 6 = c(41  95  42  88  40  30  39  24  43  32  22  69 108  96  44  94  26 100)
# Trial 7 = c(29  39  87  23  46  43  88  26  95 101  40  85  69  12  32  94  42  73)
# Trial 8 = c(87  40  88  27  95 108  69  24  43  28  30  73 100  42  26  31  46  97)
# Trial 9 = c(100  97  42  29  27  12  73 101  94  31  85  32  24  43  96 108  88  30)
# Trial 10 = c(88  31  95  12  72  39  42  28  24  30  22 101  32  26  27  69  46 100)



r18.sedge99 <- rando18.sedge99 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.sedge18.99 <- r18.sedge99[, colSums(r18.sedge99 != 0) > 0]
ncol(sp.sedge18.99) 
# Trial 1 = 30
# Trial 2 = 30
# Trial 3 = 33
# Trial 4 = 32
# Trial 5 = 32
# Trial 6 = 31
# Trial 7 = 34
# Trial 8 = 31
# Trial 9 = 31
# Trial 10 = 32


# alpha diversity
pa.sedge18.99 <- apply(sp.sedge18.99 >0,1,sum) # convert cover value to presence/absence
mean(pa.sedge18.99) 

# Trial 1 = 7.67
# Trial 2 = 8.56
# Trial 3 = 8.56
# Trial 4 = 8.33
# Trial 5 = 8.06
# Trial 6 = 8.06
# Trial 7 = 8.72
# Trial 8 = 8.33
# Trial 9 = 8.56
# Trial 10 = 8.22


sd(pa.sedge18.99) 
# Trial 1 = 1.71
# Trial 2 = 2.06
# Trial 3 = 2.06
# Trial 4 = 1.78
# Trial 5 = 1.89
# Trial 6 = 2.10
# Trial 7 = 1.93
# Trial 8 = 2.22
# Trial 9 = 1.95
# Trial 10 = 2.05


# beta diversity
ncol(sp.sedge18.99)/mean(pa.sedge18.99) 
# Trial 1 = 3.91
# Trial 2 = 3.51
# Trial 3 = 3.86
# Trial 4 = 3.84
# Trial 5 = 3.97
# Trial 6 = 3.85
# Trial 7 = 3.90
# Trial 8 = 3.72
# Trial 9 = 3.62
# Trial 10 = 3.89


##~~~~~~~~~~~
# Fescue
##~~~~~~~~~~~

fescue.99 <- grps.99 %>% 
  filter(Assemblage == "fescue") %>% 
  select(-Assemblage, -Year)

#select 18 random plots
rando18.fescue99 <- fescue.99[sample(nrow(fescue.99), 18), ]
unique(rando18.fescue99$Quadrat)
# Trial 1 = c(56  62  66  65  11  47  55  64  60  67  10  76  48  59  49  57  20 107)
# Trial 2 = c(74 106  58  64  11  63  57  55  61  67  48  75  60  62  21  68  59  25)
# Trial 3 = c(10 48 62 57 56 60 65 58 59 63  9 86 11 38 49 25 64 76)
# Trial 4 = c(21  11  61  75 106  50  93  20  66  62  60  37  47  86  38  49  55   9)
# Trial 5 = c(65  58 107  76  60  62  48  20  47  10  56  61  64  68  37  25  66  57)
# Trial 6 = c(38  10  25  47  21 106  50  64  86  66 107  76  48  63  57  62  61  75)
# Trial 7 = c(65 106  93  55  76  50  10  37  64 107  63  66  56  21  61   9  74  59)
# Trial 8 = c(48  38  11  67  56  64  74  65  59  76 106  25  47  10  93  55  58  60)
# Trial 9 = c(50 76 47 75 66 57 38 64 65 11 10 59 58 56 25  9 86 49)
# Trial 10 = c(25 74 10 64 37 93 58 76 86  9 62 57 65 55 50 38 56 47)


r18.fescue99 <- rando18.fescue99 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.fescue18.99 <- r18.fescue99[, colSums(r18.fescue99 != 0) > 0]
ncol(sp.fescue18.99) 

# Trial 1 = 35
# Trial 2 = 37
# Trial 3 = 36
# Trial 4 = 36
# Trial 5 = 36
# Trial 6 = 37
# Trial 7 = 36
# Trial 8 = 38
# Trial 9 = 36
# Trial 10 = 37

# alpha diversity
pa.fescue18.99 <- apply(sp.fescue18.99 >0,1,sum) # convert cover value to presence/absence
mean(pa.fescue18.99) 

# Trial 1 = 10.61
# Trial 2 = 10.28
# Trial 3 = 9.83
# Trial 4 = 9.22
# Trial 5 = 10.67
# Trial 6 = 9.83
# Trial 7 = 8.89
# Trial 8 = 9.28
# Trial 9 = 9.5
# Trial 10 = 9.11


sd(pa.fescue18.99) 

# Trial 1 = 3.60
# Trial 2 = 3.82
# Trial 3 = 3.79
# Trial 4 = 3.93
# Trial 5 = 4.10
# Trial 6 = 4.23
# Trial 7 = 3.79
# Trial 8 = 4.17
# Trial 9 = 3.84
# Trial 10 = 3.92


# beta diversity
ncol(sp.fescue18.99)/mean(pa.fescue18.99) 

# Trial 1 = 3.30
# Trial 2 = 3.6
# Trial 3 = 3.66
# Trial 4 = 3.90
# Trial 5 = 3.38
# Trial 6 = 3.76
# Trial 7 = 4.05
# Trial 8 = 4.10
# Trial 9 = 3.79
# Trial 10 = 4.06

##~~~~~~~~~~~
# Bogbean
##~~~~~~~~~~~

###** All trials will have same 18 plots; not recording all 10x 

bogbean.99 <- grps.99 %>% 
  filter(Assemblage == "bogbean") %>% 
  select(-Assemblage, -Year)

#select 18 random plots
rando18.bogbean99 <- bogbean.99[sample(nrow(bogbean.99), 18), ]
unique(rando18.bogbean99$Quadrat)

# Trial 1 = c(82  99  79  83 103 104  14  81  77  84  15  78  16  13 102  98 105  80)
# Trial 2 = c(14 104 102  81  77 105  98  16  15  78  82  80  79  84  13  99 103  83)
# Trial 3 = 
# Trial 4 = 
# Trial 5 = 
# Trial 6 = 
# Trial 7 = 
# Trial 8 =
# Trial 9 = 
# Trial 10 = 


r18.bogbean99 <- rando18.bogbean99 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.bogbean18.99 <- r18.bogbean99[, colSums(r18.bogbean99 != 0) > 0]
ncol(sp.bogbean18.99) 

# Trial 1 = 36
# Trial 2 = 36
# Trial 3 = 
# Trial 4 = 
# Trial 5 = 
# Trial 6 = 
# Trial 7 = 
# Trial 8 =
# Trial 9 = 
# Trial 10 = 

# alpha diversity
pa.bogbean18.99 <- apply(sp.bogbean18.99 >0,1,sum) # convert cover value to presence/absence
mean(pa.bogbean18.99) 

# Trial 1 = 11.5
# Trial 2 = 11.5
# Trial 3 = 
# Trial 4 = 
# Trial 5 = 
# Trial 6 = 
# Trial 7 = 
# Trial 8 =
# Trial 9 = 
# Trial 10 = 


sd(pa.bogbean18.99) 

# Trial 1 = 2.92
# Trial 2 = 2.92
# Trial 3 = 
# Trial 4 = 
# Trial 5 = 
# Trial 6 = 
# Trial 7 = 
# Trial 8 =
# Trial 9 = 
# Trial 10 = 


# beta diversity
ncol(sp.bogbean18.99)/mean(pa.bogbean18.99) 

# Trial 1 = 3.13
# Trial 2 = 3.13
# Trial 3 = 
# Trial 4 = 
# Trial 5 = 
# Trial 6 = 
# Trial 7 = 
# Trial 8 =
# Trial 9 = 
# Trial 10 = 


##~~~~~~~~~~~
# Total
##~~~~~~~~~~~

rando18.99 <- lst(rando18.sedge99, rando18.fescue99, rando18.bogbean99) %>% 
  reduce(full_join) %>% 
  replace(is.na(.), 0) %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.rand18.99 <- rando18.99[, colSums(rando18.99 != 0) > 0]
ncol(sp.rand18.99) 

# Trial 1 = 42
# Trial 2 = 41
# Trial 3 = 43
# Trial 4 = 36
# Trial 5 = 42
# Trial 6 = 43
# Trial 7 = 44
# Trial 8 = 44
# Trial 9 = 43
# Trial 10 = 43



# alpha diversity
pa18.99 <- apply(sp.rand18.99 >0,1,sum) # convert cover value to presence/absence
mean(pa18.99) 

# Trial 1 = 9.93
# Trial 2 = 10.11
# Trial 3 =  9.96
# Trial 4 = 11.5
# Trial 5 = 10.07
# Trial 6 = 9.80
# Trial 7 = 9.70
# Trial 8 = 9.70
# Trial 9 = 9.85
# Trial 10 = 9.61


sd(pa18.99) 

# Trial 1 = 3.25
# Trial 2 = 3.20
# Trial 3 = 3.19
# Trial 4 = 3.92
# Trial 5 = 3.39
# Trial 6 = 3.45
# Trial 7 = 3.19
# Trial 8 = 3.42
# Trial 9 = 3.19
# Trial 10 = 3.31

# beta diversity
ncol(sp.rand18.99)/mean(pa18.99) 

# Trial 1 = 4.23
# Trial 2 = 4.05
# Trial 3 = 4.32
# Trial 4 = 3.13
# Trial 5 = 4.17
# Trial 6 = 4.39
# Trial 7 = 4.53
# Trial 8 = 4.53
# Trial 9 = 4.36
# Trial 10 = 4.47

```

```{r bootstrap 18 plots - 1979}

# grps.79 <-  d1979_grps %>% 
#   mutate(Assemblage = ifelse(Association_Type2 == 1, "sedge",
#                       ifelse(Association_Type2 == 2, "fescue", "bogbean"))) %>% 
#   relocate(Assemblage, .after = Year) %>% 
#   select(-Transect, -Association_Type2) 

# select 18 random plots from each assemblage type & calculate diversity components

##~~~~~~~~~~~
# Sedge
##~~~~~~~~~~~
# sedge.79 <- grps.79 %>% 
#   filter(Assemblage == "sedge") %>% 
#   select(-Assemblage, -Year)

#select 18 random plots
rando18.sedge79 <- sedge.79[sample(nrow(sedge.79), 18), ]
unique(rando18.sedge79$Quadrat)

# Trial 1 = c(32  49 108  23  24  93  28  97  20  56  96  73  87  30  31  72  68  22)
# Trial 2 = c(49 29 23 57 69 26 22 93 86 25 87 30 94 28 97 20 27 48)
# Trial 3 = c(57  27  31  58  95  97  48  87  28  86  88  25 108  94  96  22  69  24)
# Trial 4 = c(94  49  27  24  23  29  86  32 108  73  56  97  48  25  26  28  47  69)
# Trial 5 = c(47  88  20  94  29  28  60  32  23  95  87  73  55  48  25 108  86  22)
# Trial 6 = c(108  55  25  97  94  24  49  31  29  47  57  48  59  73  20  87  96  72)
# Trial 7 = c(58 31 73 20 72 96 87 24 86 27 23 30 29 57 60 69 93 94)
# Trial 8 = c(97  95  20 108  94  59  27  47  48  69  22  49  72  86  58  68  26  25)
# Trial 9 = c(86 30 23 47 96 94 26 68 22 49 56 69 57 20 72 59 93 58)
# Trial 10 = c(23 26 93 48 88 32 30 69 27 24 31 59 29 87 28 72 57 25)



r18.sedge79 <- rando18.sedge79 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.sedge18.79 <- r18.sedge79[, colSums(r18.sedge79 != 0) > 0]
ncol(sp.sedge18.79) 
# Trial 1 =  31
# Trial 2 = 33
# Trial 3 = 29
# Trial 4 = 32
# Trial 5 = 34
# Trial 6 = 32
# Trial 7 = 35
# Trial 8 = 34
# Trial 9 = 34
# Trial 10 = 29


# alpha diversity
pa.sedge18.79 <- apply(sp.sedge18.79 >0,1,sum) # convert cover value to presence/absence
mean(pa.sedge18.79) 

# Trial 1 = 10.28
# Trial 2 = 10.78
# Trial 3 = 10.28
# Trial 4 = 10.83
# Trial 5 = 10.22
# Trial 6 = 10.67
# Trial 7 = 11.22
# Trial 8 = 11.0
# Trial 9 = 11.39
# Trial 10 = 10.06


sd(pa.sedge18.79) 
# Trial 1 = 2.42
# Trial 2 = 2.32
# Trial 3 = 2.27
# Trial 4 = 2.62
# Trial 5 = 2.71
# Trial 6 = 2.28
# Trial 7 = 2.32
# Trial 8 = 2.40
# Trial 9 = 2.12
# Trial 10 = 1.89


# beta diversity
ncol(sp.sedge18.79)/mean(pa.sedge18.79) 
# Trial 1 = 3.02
# Trial 2 = 3.06
# Trial 3 = 2.82
# Trial 4 = 2.95
# Trial 5 = 3.33
# Trial 6 = 3.0
# Trial 7 = 3.12
# Trial 8 = 3.09
# Trial 9 = 2.99
# Trial 10 = 2.88


##~~~~~~~~~~~
# Fescue
##~~~~~~~~~~~
# 
# fescue.79 <- grps.79 %>% 
#   filter(Assemblage == "fescue") %>% 
#   select(-Assemblage, -Year)

#select 18 random plots
rando18.fescue79 <- fescue.79[sample(nrow(fescue.79), 18), ]
unique(rando18.fescue79$Quadrat)
# Trial 1 = c(74  39  77  16  40  42  11 106 104  76 103  10 102 105  44  43  37 107)
# Trial 2 = c(40  13  75  11   9  43 105 104 106 103  77  44  37  76  14  39  46 102)
# Trial 3 = c(43  46  40  38  14 107 104  37  77 105  39  75 106  15  11  74  16 103)
# Trial 4 = c(40  75  14 105  15  41  42  38 107 106   9  77 102 101  10  74  13  43)
# Trial 5 = c(44  41 103   9 104  40  37  43  11  14  38  39 106  42  13  15 101 102)
# Trial 6 = c(43  44  13  16 102  46  39  76 101  42 107  15  41  75  37  21   9  10)
# Trial 7 = c(74  15  77  11 106  16  39 102  75 103 107  76 104  44 105  21 101  46)
# Trial 8 = c(9  44  75  40 103  76  41 102 104  12  10  21  15  42  13  11  77 105)
# Trial 9 = c(105  10  74  76   9  13  15 103  44  16  75  43  11  14  41  38  46 101)
# Trial 10 = c(15  75 107   9  13  11  16  41  38 105  77  42  43  37 101  46 103  21)


r18.fescue79 <- rando18.fescue79 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.fescue18.79 <- r18.fescue79[, colSums(r18.fescue79 != 0) > 0]
ncol(sp.fescue18.79) 

# Trial 1 = 44
# Trial 2 = 44
# Trial 3 = 43
# Trial 4 = 43
# Trial 5 = 43
# Trial 6 = 41
# Trial 7 = 44
# Trial 8 = 45
# Trial 9 = 43
# Trial 10 = 43

# alpha diversity
pa.fescue18.79 <- apply(sp.fescue18.79 >0,1,sum) # convert cover value to presence/absence
mean(pa.fescue18.79) 

# Trial 1 = 13.67
# Trial 2 = 13.61
# Trial 3 = 12.89
# Trial 4 = 12.0
# Trial 5 = 13.44
# Trial 6 = 12.72
# Trial 7 = 13.72
# Trial 8 = 13.28
# Trial 9 = 12.11
# Trial 10 = 12.17


sd(pa.fescue18.79) 

# Trial 1 = 4.35
# Trial 2 = 4.10
# Trial 3 = 3.88
# Trial 4 = 3.55
# Trial 5 = 4.27
# Trial 6 = 3.82
# Trial 7 = 3.79
# Trial 8 = 3.98
# Trial 9 = 3.92
# Trial 10 = 3.42


# beta diversity
ncol(sp.fescue18.79)/mean(pa.fescue18.79) 

# Trial 1 = 3.22
# Trial 2 = 3.23
# Trial 3 = 3.34
# Trial 4 = 3.58
# Trial 5 = 3.20
# Trial 6 = 3.22
# Trial 7 = 3.21
# Trial 8 = 3.39
# Trial 9 = 3.55
# Trial 10 = 3.53

##~~~~~~~~~~~
# Bogbean
##~~~~~~~~~~~

# bogbean.79 <- grps.79 %>% 
#   filter(Assemblage == "bogbean") %>% 
#   select(-Assemblage, -Year)

#select 18 random plots
rando18.bogbean79 <- bogbean.79[sample(nrow(bogbean.79), 18), ]
unique(rando18.bogbean79$Quadrat)

# Trial 1 = c(61  63  64  79 100  78  81  67  98  80  85  62  82  84  66  83  99  50)
# Trial 2 = c(78  83  61  62  65  63  98  84  99  64  81 100  85  79  80  50  67  82)
# Trial 3 = c(50  99  79  63  62 100  81  84  66  78  83  98  65  61  67  80  82  85)
# Trial 4 = c(78  85  67  62  98  65  84  81  83  79  80  63  61  82  64 100  50  66)
# Trial 5 = c(84  83  63  81  66  99  80  61  64  79  82  65  85 100  62  50  98  67)
# Trial 6 = c(85  62  50  66  83  99  98  82  67  81  80 100  79  63  78  65  64  61)
# Trial 7 = c(85  79  67  61  63  78  83  80  66  84  99  81  65 100  64  82  62  50)
# Trial 8 = c(82  66  62  67  80  98  61  99 100  83  63  50  65  84  64  85  78  79)
# Trial 9 = c(100  84  85  83  99  67  80  62  78  79  63  66  64  65  61  82  98  81)
# Trial 10 = c(79  64  83  67  99  66  82  78  85 100  50  63  98  62  80  61  84  81)


r18.bogbean79 <- rando18.bogbean79 %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.bogbean18.79 <- r18.bogbean79[, colSums(r18.bogbean79 != 0) > 0]
ncol(sp.bogbean18.79) 

# Trial 1 = 31
# Trial 2 = 32
# Trial 3 = 32
# Trial 4 = 32
# Trial 5 = 31
# Trial 6 = 32
# Trial 7 = 32
# Trial 8 = 32
# Trial 9 = 32
# Trial 10 = 31

# alpha diversity
pa.bogbean18.79 <- apply(sp.bogbean18.79 >0,1,sum) # convert cover value to presence/absence
mean(pa.bogbean18.79) 

# Trial 1 = 12.61
# Trial 2 = 12.89
# Trial 3 = 13.06
# Trial 4 = 13.0
# Trial 5 = 12.56
# Trial 6 = 12.94
# Trial 7 = 13.0
# Trial 8 = 12.72
# Trial 9 = 12.89
# Trial 10 = 12.61


sd(pa.bogbean18.79) 

# Trial 1 = 3.57
# Trial 2 = 3.71
# Trial 3 = 3.59
# Trial 4 = 3.65
# Trial 5 = 3.48
# Trial 6 = 3.69
# Trial 7 = 3.65
# Trial 8 = 3.68
# Trial 9 = 3.71
# Trial 10 = 3.57


# beta diversity
ncol(sp.bogbean18.79)/mean(pa.bogbean18.79) 

# Trial 1 = 2.46
# Trial 2 = 2.48
# Trial 3 = 2.45
# Trial 4 = 2.46
# Trial 5 = 2.47
# Trial 6 = 2.47
# Trial 7 = 2.46
# Trial 8 = 2.52
# Trial 9 = 2.48
# Trial 10 = 2.46


##~~~~~~~~~~~
# Total
##~~~~~~~~~~~

rando18.79 <- lst(rando18.sedge79, rando18.fescue79, rando18.bogbean79) %>% 
  reduce(full_join) %>% 
  replace(is.na(.), 0) %>% 
  select(-Quadrat)

# total number of species = ncol > 0
sp.rand18.79 <- rando18.79[, colSums(rando18.79 != 0) > 0]
ncol(sp.rand18.79) 

# Trial 1 = 47
# Trial 2 = 48
# Trial 3 = 47
# Trial 4 = 49
# Trial 5 = 47
# Trial 6 = 47
# Trial 7 = 48
# Trial 8 = 49
# Trial 9 = 48
# Trial 10 = 46



# alpha diversity
pa18.79 <- apply(sp.rand18.79 >0,1,sum) # convert cover value to presence/absence
mean(pa18.79) 

# Trial 1 = 12.19
# Trial 2 = 12.43
# Trial 3 = 12.07
# Trial 4 = 11.94
# Trial 5 = 12.07
# Trial 6 = 12.11
# Trial 7 = 12.65
# Trial 8 = 12.33
# Trial 9 = 12.13
# Trial 10 = 11.61


sd(pa18.79) 

# Trial 1 = 3.75
# Trial 2 = 3.61
# Trial 3 = 3.50
# Trial 4 = 3.36
# Trial 5 = 3.74
# Trial 6 = 3.43
# Trial 7 = 3.42
# Trial 8 = 3.50
# Trial 9 = 3.34
# Trial 10 = 3.20

# beta diversity
ncol(sp.rand18.79)/mean(pa18.79) 

# Trial 1 = 3.86
# Trial 2 = 3.86
# Trial 3 = 3.89
# Trial 4 = 4.10
# Trial 5 = 3.89
# Trial 6 = 3.88
# Trial 7 = 3.80
# Trial 8 = 3.97
# Trial 9 = 3.96
# Trial 10 = 3.96

```


Stability indices using codyn
```{r community_stability}
# library(codyn)
## see https://rdrr.io/cran/codyn/man/community_stability.html
## see https://cran.r-project.org/web/packages/codyn/vignettes/Temporal_Diversity_Indices.html 

join_40 <- lst(d1979_grps, d1999_grps, d2019_grps) %>% 
  reduce(full_join) %>% 
  replace(is.na(.), 0) %>% 
  pivot_longer(AGST2:SYSU4, names_to = "Species", values_to = "Cover_class") %>% 
   mutate(Assemblage = ifelse(Association_Type2 == 1, "Sedge",
                      ifelse(Association_Type2 == 2, "Fescue", "Bogbean"))) %>% 
  select(-Transect, -Association_Type2, -Association_Type, -Association_Subtype) %>% 
  group_by(Year, Assemblage, Species) %>% 
  summarize(mean_cover = mean(Cover_class, na.rm = TRUE))
 

# measure of stability
cs_avg <- community_stability(join_40, time.var = "Year", abundance.var = "mean_cover")
# avg. stability = 4.282


cs <- community_stability(join_40, time.var = "Year", abundance.var = "mean_cover", replicate.var = "Assemblage")
kable(cs)
# 
# Assemblage  stability
# Bogbean	    7.173355
# Fescue	    2.521427			
# Sedge	      5.082942

# test of significance: variance ratio
# set bootnumber lower for faster processing; for final use set higher (eg, 9999)
res_averagedreplicates <- variance_ratio(join_40,
               time.var = "Year",
               species.var = "Species",
               abundance.var = "mean_cover",
               bootnumber = 999,
               replicate = "Assemblage")
# Avg VR = 3.22

res_assemblage <- variance_ratio(join_40,
               time.var = "Year",
               species.var = "Species",
               abundance.var = "mean_cover",
               bootnumber = 999,
               replicate = "Assemblage",
               average.replicates = "FALSE")
# Bogbean VR = 1.01
# Fescue VR = 6.16
# Sedge VR = 2.39
## VR = 1 indicates species do not covary; > 1 indicates positive covariance among species. 

cs_avg <- data.frame(Assemblage = "Average", stability = cs_avg, lowerCI = res_averagedreplicates$lowerCI, upperCI = res_averagedreplicates$upperCI, nullmean = res_averagedreplicates$nullmean, VR = res_averagedreplicates$VR)

cs_vr <- cs %>% 
  inner_join(res_assemblage, by = "Assemblage") %>% 
  rbind(cs_avg)              

cs_vr$Assemblage <- factor(cs_vr$Assemblage, levels = c("Bogbean", "Fescue", "Sedge", "Average"))
ggplot(cs_vr, aes(VR, stability, color = Assemblage, shape = Assemblage)) + 
   geom_point(size = 6) + theme_classic() +
   theme(text = element_text(size = 18)) +  
   labs(x = "variance ratio", 
        y = "stability 
        (temporal mean/temporal sd)") +
   scale_color_manual(values = c("#0072B2", "#E69F00", "#009E73", "#000000"))

# See Fig. 3 at https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.12569
## High variance ratio should indicate low stability, which is reflected in data here. 
## Bogbean assemblage has highest stability, and lowest VR, while Fescue group has greatest VR and thus lowest stability. Sedge is intermediate

```

```{r rank clock plots}
library(wesanderson)

# see https://cran.r-project.org/web/packages/codyn/vignettes/Temporal_Diversity_Indices.html 
#need to select species of interest - use indicator analysis? rare/invasive/foundaitonal?
## indicator analysis that appear in all 3 years? Poa palustris not in 2019...
############################################################################################################
# all species, but rainbow legend is hard to interpret

# may need to use join.grps
clock_sp <- join_40 %>% 
   filter(Species %in% c("AGST2", "BICE", "CALY3", "DECE", "ELPA3", "EQFL", "EQPA", "EQVA", "FEAR3", 
                         "GATR3", "IMCA", "JUAR",  "LAPA4", "LYSA2", "LYTH2", "MEAQ", "METR3", "MYSC", 
                         "PHAR3", "POPA2", "SALA2", "SALUL", "SCTA2", "SIHE4", "TRWO"))
                         
 
# ggplot(clock_sp, aes(Year, mean_cover, color = Species)) + 
#    geom_line(size = 2) + coord_polar() + theme_classic() +
#    facet_wrap(~Assemblage) 
 
 
 # select only 2 most significant species from each year
# clock_select_sp <- join_40 %>% 
#    filter(Species %in% c("CALY3", "SALA2", "AGST2", "FEAR3", "SALUL", "POPA2", "MEAQ", "METR3", "MYSC", "PHAR3"))
#             
# ggplot(clock_select_sp, aes(Year, mean_cover, color = Species)) + 
#    geom_line(size = 2) + coord_polar() + theme_classic() +
#    facet_wrap(~Assemblage) + 
#    labs(x = "", 
#         y = "Cover abundance class (mean)")
############################################################################################################
# By assemblage. Clunky, but easier to manipulate legend and read
 #Note: Species shown are the two most significant from the indicator species analysis. 
#########################################
# Bogbean
#########################################

clock_bogbean <- join_40 %>% 
   filter(Assemblage == "Bogbean") %>% 
   filter(Species %in% c("METR3", "MYSC", "BICE", "LYSA2", "EQFL", "LYTH2", "MEAQ", "JUAR", "ELPA3",  "DECE")) %>% 
   mutate(Species = recode(Species, "METR3" = "Menyanthes trifoliata", "MYSC" = "Myosotis scorpiodes", "BICE" = "Bidens cernua", "LYSA2" = "Lythrum salicaria", 
                            "EQFL" = "Equisetum fluviatile", "LYTH2" = "Lysimachia thyrsiflora", "MEAQ" = "Mentha aquatica", "JUAR" = "Juncus articulatus", 
                           "ELPA3" = "Eleocharis palustris", "DECE" = "Deschampsia caespitosa")) 

# select only three most significant species per year
clock_bogbean_select <- join_40 %>% 
   filter(Assemblage == "Bogbean") %>% 
   filter(Species %in% c("METR3", "MYSC", "MEAQ", "BICE", "LYSA2", "LYTH2")) %>% 
   mutate(Species = recode(Species, "METR3" = "Menyanthes trifoliata", "MYSC" = "Myosotis scorpiodes", "MEAQ" = "Mentha aquatica", "BICE" = "Bidens cernua", "LYSA2" = "Lythrum salicaria", "LYTH2" = "Lysimachia thyrsiflora"))
 
bogbean.clock <- ggplot(clock_bogbean_select, aes(Year, mean_cover, color = Species)) + 
   geom_line(size = 2) + coord_polar() + theme_classic() +
   theme(text = element_text(size = 18), 
         #axis.text.x=element_blank(), # remove all x axis labels in polar_coord
         #axis.text.x=element_text(angle = 25),
         legend.text = element_text(face = "italic"), 
         legend.title=element_blank(), 
         legend.position="right") +
   #guides(colour = guide_legend(nrow = 3)) + 
  #geom_text(aes(x = Year + 0.5,label = Year)) +
  scale_x_continuous(breaks=c(1981, 1999, 2018), labels = c(1979, 1999, 2019)) +
  #guides(x = guide_axis(angle = 90)) +
  #guides(x = guide_axis(n.dodge = 5)) +
  labs(x = "Bogbean", y = "") +
   scale_color_brewer(palette = "Dark2")
bogbean.clock



#########################################
# Fescue
#########################################
# elected to keep top two indicator species for visualization due to greatest changes observed

clock_fescue <- join_40 %>% 
   filter(Assemblage == "Fescue") %>% 
   filter(Species %in% c("FEAR3", "SALUL",  "SIHE4", "POPA2", "PHAR3",  "AGST2")) %>% 
    mutate(Species = recode(Species, "FEAR3" = "Festuca arundinacea", "SALUL" = "Salix lasiandra",  "SIHE4" = "Sidalcia hendersonii", "POPA2" = "Poa palustris", "PHAR3" = "Phalaris arundinacea", "AGST2" = "Agrostis stolonifera")) 

fescue.clock <- ggplot(clock_fescue, aes(Year, mean_cover, color = Species)) + 
   geom_line(size = 2) + coord_polar() + theme_classic() +
     theme(text = element_text(size = 18), 
         legend.text = element_text(face = "italic"), 
         legend.title=element_blank(), 
         legend.position="right") +
   #guides(colour = guide_legend(nrow = 3)) + 
  scale_x_continuous(breaks=c(1981, 1999, 2018), labels = c(1979, 1999, 2019)) +
  labs(x = "Fescue", 
        y = "", 
        color = "Species") +
   scale_color_brewer(palette = "Set2")

#########################################
 # Sedge
#########################################
clock_sedge <- join_40 %>% 
   filter(Assemblage == "Sedge") %>% 
   filter(Species %in% c("CALY3", "SALA2", "SCTA2", "AGST2", "IMCA")) %>% 
    mutate(Species = recode(Species, "CALY3" = "Carex lyngbyei", "SALA2" = "Sagittaria latifolia", "SCTA2" = "Schoenoplectus tabernaemontani",
                           "AGST2" = "Agrostis stolonifera", "IMCA" = "Impatiens capensis"))

sedge.clock<- ggplot(clock_sedge, aes(Year, mean_cover, color = Species)) + 
   geom_line(size = 2) + coord_polar() + theme_classic() +
     theme(text = element_text(size = 18), 
         legend.text = element_text(face = "italic"), 
         legend.title=element_blank(), 
         legend.position="right") +
   #guides(colour = guide_legend(nrow = 3)) + 
  scale_x_continuous(breaks=c(1981, 1999, 2018), labels = c(1979, 1999, 2019)) +
    labs(x = "Sedge", 
        y = "", 
        color = "Species") +
   scale_color_manual(values = wes_palette("Moonrise3"))

## panel format w/ ggpubr
library(ggpubr)
clock.panel <- ggarrange(sedge.clock, fescue.clock, bogbean.clock,
          ncol = 1, nrow = 3, align = "hv") 
clock.panel

clock.ann <- annotate_figure(clock.panel, left = text_grob("Cover abundance class (mean)", size = 20, rot = 90, hjust = .5, vjust = 1))
clock.ann

#ggsave("fig-clock-panel.jpg", clock.ann, width = 12, height = 16)

```

DELETE or archive below this line once final analyses are done
```{r NMDS w/ 95% CI, ANOSIM, PERMANOVA}

# largely based on https://stackoverflow.com/questions/13794419/plotting-ordiellipse-function-from-vegan-package-onto-nmds-plot-created-in-ggplo

# see also
# http://kembellab.ca/r-workshop/biodivR/SK_Biodiversity_R.html
# https://stackoverflow.com/questions/46132682/overlap-percentage-of-nmds-using-metamds-in-vegan-package-in-r

join_40 <- lst(d1979_grps, d1999_grps, d2019_grps) %>% 
  reduce(full_join) %>% replace(is.na(.), 0) #%>% 
  #select(Elev_AMSL_m:SLL_Quadrat, everything())
nrow(join_40) #231


# two outliers: 2019 site 87 is an outlier on NMDS plot. remove here
join_40_no_outliers <- join_40[!(join_40$Year == 2019 & join_40$Quadrat == 87),]
#nrow(join_40_no_outliers) #230
  
join_40_no_outliers$Association_Type2 <- factor(join_40_no_outliers$Association_Type2, levels = c("1", "2", "3"),
                  labels = c("Sedge", "Fescue", "Bogbean"))

# rerun NMDS w/ outliers removed. All descriptive text is removed here.

com_out = join_40_no_outliers[,11:ncol(join_40_no_outliers)]
m_com_out = as.matrix(com_out)
set.seed(123)
nmds_out = metaMDS(m_com_out, distance = "bray")
nmds_out # stress = 0.232

data.scores_out = as.data.frame(scores(nmds_out))

data.scores_out$Year = join_40_no_outliers$Year
data.scores_out$Association_Type2 = join_40_no_outliers$Association_Type2

Ladner.scores_out <- data.scores_out %>%
  mutate(Year = as.character(Year)) %>%
           mutate(Association_Type2 = as.character(Association_Type2))

#distance for NMDS
sol <- metaMDS(com_out, distance = "bray", k = 3, trymax = 100) 
# plot(sol$points, col = join_40_no_outliers$Year)
# ordiellipse(sol, join_40_no_outliers$Year, display = "sites", kind = "sd", label = T)


NMDS = data.frame(MDS1 = sol$points[,1], MDS2 = sol$points[,2], Year=join_40_no_outliers$Year, Comm = join_40_no_outliers$Association_Type2)
NMDS$Comm <- as.factor(NMDS$Comm)
NMDS$Year <- as.factor(NMDS$Year)

#NMDS.mean = aggregate(NMDS[,1:2],list(Year=NMDS$Year),mean)
ggplot(NMDS, aes(x = MDS1, y = MDS2)) + 
  geom_point(aes(shape = Year, color = Year))


veganCovEllipse<-function (cov, center = c(0, 0), scale = 1, npoints = 100) 
{
theta <- (0:npoints) * 2 * pi/npoints
Circle <- cbind(cos(theta), sin(theta))
t(center + scale * t(Circle %*% chol(cov)))
}

df_ell <- data.frame() 

for(g in levels(NMDS$Year)){
df_ell <- rbind(df_ell, 
                cbind(as.data.frame(with(NMDS[NMDS$Year==g,],
                veganCovEllipse(cov.wt(cbind(MDS1,MDS2),
                wt=rep(1/length(MDS1),
                length(MDS1)))$cov,
                center=c(mean(MDS1),
                mean(MDS2))))),
                Year=g))}

#### 
LadnerNMDS <- ggplot(Ladner.scores_out, aes(x = NMDS1, y = NMDS2)) + 
    geom_point(size = 5, alpha = 0.7, aes(shape = Association_Type2, colour = Year))+ 
    theme(axis.text.y = element_text(colour = "black", size = 16, face = "bold"), 
    axis.text.x = element_text(colour = "black", face = "bold", size = 16), 
    legend.text = element_text(size = 14, face ="bold", colour ="black"), 
    legend.position = "bottom", axis.title.y = element_text(face = "bold", size = 14), 
    axis.title.x = element_text(face = "bold", size = 16, colour = "black"), 
    legend.title = element_text(size = 14, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank()) + 
    labs(x = "NMDS1", colour = "Observation Year", y = "NMDS2", shape = "Community Association Type", caption = "Ellipses are 95%   confidence intervals") + 
    scale_colour_manual(values = c("#4B0082", "#9370DB", "#DDA0DD")) +
    geom_path(data=df_ell, aes(x=MDS1, y=MDS2, color = Year), size =2, alpha = 0.7) 
LadnerNMDS

#ggsave("Fig_NMDS.svg", width = 10, height = 6, units = "in", device = 'svg')

################# 
# NMDS Year only (no assemblage)
NMDS_yr <- ggplot(Ladner.scores_out, aes(x = NMDS1, y = NMDS2)) + 
    geom_point(size = 5, alpha = 0.7, aes(shape = Year, colour = Year))+ 
    theme(axis.text.y = element_text(colour = "black", size = 16, face = "bold"), 
    axis.text.x = element_text(colour = "black", face = "bold", size = 16), 
    legend.text = element_text(size = 14, face ="bold", colour ="black"), 
    legend.position = "bottom", axis.title.y = element_text(face = "bold", size = 14), 
    axis.title.x = element_text(face = "bold", size = 16, colour = "black"), 
    legend.title = element_text(size = 14, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank()) + 
    labs(x = "NMDS1", y = "NMDS2", caption = "Ellipses are 95%   confidence intervals") + 
    scale_colour_manual(values = c("#4B0082", "#9370DB", "#DDA0DD")) +
    geom_path(data=df_ell, aes(x=MDS1, y=MDS2, color = Year), size =2, alpha = 0.7) 
NMDS_yr

#######################
# Analysis of Similarity
#######################
#note: ANOSIM is sensitive to inequal compositional variance (see Anderson & Walsh, 2013)
# these attempts use full years' dataset, not by community

#analysis of similarity between years of observation
ano_out = anosim(m_com_out, join_40_no_outliers$Year, distance = "bray", permutations = 9999)
ano_out
# ANOSIM statistic R: 0.2021
#       Significance: 1e-04 

#analysis of similarity between community types, combined across all years
ano_assoc_out = anosim(m_com_out, join_40_no_outliers$Association_Type2, distance = "bray", permutations = 9999)
ano_assoc_out


#######################
# PERMANOVA
#######################
permanova_out = adonis(m_com_out ~ join_40_no_outliers$Year, permutations = 999, method = "bray")


#                        Df SumsOfSqs MeanSqs F.Model      R2 Pr(>F)    
# join_40_no_outliers$Year   1     5.883  5.8833   22.25 0.08891  0.001 ***
# Residuals                228    60.287  0.2644         0.91109           
# Total                    229    66.170                 1.00000  
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# https://www.researchgate.net/post/Posthoc_test_for_permanova_adonis
#install.packages("devtools")


```

```{r NMDS, just 2019 w/ enviro loadings}
# see tutorial https://jkzorz.github.io/2020/04/04/NMDS-extras.html 
library(vegan)
library(ggplot2)

# using d2019_grps since updated SLL data w/ elev & dist from channel

com_2019 = d2019_grps[,8:50]
env_2019 = d2019_grps[,3:4]

#convert community to a matrix
m_com_2019 = as.matrix(com_2019)

#nmds code
set.seed(123)
nmds_2019 = metaMDS(m_com_2019, distance = "bray")
nmds_2019
# stress = 0.199

en_2019 = envfit(nmds_2019, env_2019, permutations = 999, na.rm = TRUE)

# basic nmds plot
plot(nmds_2019)
plot(en_2019)

# nicer plot

# extract NMDS & envfit output to vectors
scores_2019 = as.data.frame(scores(nmds_2019))
en_coord = as.data.frame(scores(en_2019, "vectors")) * ordiArrowMul(en_2019)
rownames(en_coord) <- c("Elevation AMSL (m)", "Distance from channel (m)")

# create CI and ellipses
NMDS_plot = data.frame(MDS1 = nmds_2019$points[,1], MDS2 = nmds_2019$points[,2], Comm_19 = d2019_grps$Association_Type2)
NMDS_plot$Comm_19 <- as.factor(NMDS_plot$Comm_19)


veganCovEllipse<-function (cov, center = c(0, 0), scale = 1, npoints = 100) 
{
theta <- (0:npoints) * 2 * pi/npoints
Circle <- cbind(cos(theta), sin(theta))
t(center + scale * t(Circle %*% chol(cov)))
}

df_ell_19 <- data.frame() 

for(g in levels(NMDS_plot$Comm_19)){
df_ell_19 <- rbind(df_ell_19, 
                cbind(as.data.frame(with(NMDS_plot[NMDS_plot$Comm_19==g,],
                veganCovEllipse(cov.wt(cbind(MDS1,MDS2),
                wt=rep(1/length(MDS1),
                length(MDS1)))$cov,
                center=c(mean(MDS1),
                mean(MDS2))))),
                Comm_19=g))}

# plot the whole thing
plot_2019 = ggplot(NMDS_plot, aes(x = MDS1, y = MDS2)) + 
    geom_segment(aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2), 
       data = en_coord, size =1, alpha = 0.5, colour = "grey30") +
    geom_text(data = en_coord, aes(x = NMDS1, y = NMDS2), colour = "black", 
       fontface = "bold", label = row.names(en_coord)) +
    geom_point(size = 5, alpha = 0.5, aes(shape = Comm_19, colour = Comm_19))+ 
    geom_path(data=df_ell_19, aes(x=MDS1, y=MDS2, color = Comm_19), size =2, alpha = 0.6) +
    theme(axis.text.y = element_text(colour = "black", size = 16, face = "bold"), 
    axis.text.x = element_text(colour = "black", face = "bold", size = 16), 
    legend.text = element_text(size = 14, face ="bold", colour ="black"), 
    legend.position = "bottom", axis.title.y = element_text(face = "bold", size = 14), 
    axis.title.x = element_text(face = "bold", size = 16, colour = "black"), 
    legend.title = element_text(size = 14, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank()) + 
    labs(x = "NMDS1", y = "NMDS2", colour = "Community Association Type", caption = "2019 Community Survey")  + 
    scale_colour_manual(values = c("#1B7837", "#FDAE61", "#762A83"), labels = c("Sedge", "Fescue", "Bogbean"))  
plot_2019

#ggsave("2019_NMDS-enviro.svg")


# library(RColorBrewer)
#brewer.pal(9, "Greys")

anosim_d2019_grps = anosim(m_com_2019, d2019_grps$Association_Type2, distance = "bray", permutations = 9999)
anosim_d2019_grps
# ANOSIM statistic R: 0.584 
#       Significance: 1e-04 

```

```{r PERMDISP & BETADISP}

###################
# PERMDISP (permutational analysis of multivariate disperson)
# dispersion tests: is compositional variance equal? that is, are groups homogeneously dispersed wrt species sampled? 
##################

# structured following:
# https://rfunctions.blogspot.com/2019/03/betadisper-and-adonis-homogeneity-of.html
# https://mattsigal.github.io/eqcov_supp/betadisp-ex.html


#####
# select community groups
#####
bogbean <- join_40_no_outliers %>% 
  filter(Association_Type2 == "Bogbean")

sedge <- join_40_no_outliers %>% 
  filter(Association_Type2 == "Sedge")

fescue <- join_40_no_outliers %>% 
  filter(Association_Type2 == "Fescue")

#####
# cover as matrix
#####
bogbean_out = bogbean[,7:ncol(bogbean)]
matrix_bogbean = as.matrix(bogbean_out)

sedge_out = sedge[,10:ncol(sedge)]
matrix_sedge = as.matrix(sedge_out)

fescue_out = fescue[,10:ncol(fescue)]
matrix_fescue = as.matrix(fescue_out)

#####
# calculate distances
#####
dst_bogbean <- vegdist(bogbean[,7:78], method = "bray")

dst_sedge <- vegdist(sedge[,10:81], method = "bray")

dst_fescue <- vegdist(fescue[,10:81], method = "bray")

#####
# calculate dispersion
#####

betadisper_bogbean <- betadisper(dst_bogbean, bogbean$Year)

betadisper_sedge <- betadisper(dst_sedge, sedge$Year)

betadisper_fescue <- betadisper(dst_fescue, fescue$Year)

#####
# use group dispersions to perform ANOVA/TukeyHSD, or permutation for CCA or RDA to test sig dif of dispersions. 
# a > 0.05 means group dispersions are homogenous ("null hypothesis of no difference in dispersion between groups")
#####


anova(betadisper_bogbean)
# F = 2.61, P = 0.08
permutest(betadisper_bogbean)
# F = 2.61, p = 0.094; years are homogeneously dispersed (compositional variance is equal)

anova(betadisper_fescue)
# f = 1.294, P = 0.28
permutest(betadisper_fescue)
#F = 1.294, p = 0.297

anova(betadisper_sedge)
# f = 3.44, P = 0.036
permutest(betadisper_sedge)
# F = 3.44, p = 0.044

#####
# plot dispersions to visualize heterogeneity
#####
plot(betadisper_bogbean)
plot(betadisper_sedge)
plot(betadisper_fescue)


###################
# PERMANOVA (permutational analysis of variance)
# adonis: are groups composed of different species? 
##################
adonis(matrix_bogbean ~ bogbean$Year, perm = 999, method = "bray")

adonis(matrix_sedge ~ sedge$Year, perm = 999, method = "bray")

adonis(matrix_fescue ~ fescue$Year, perm = 999, method = "bray")


######
# boxplot dispersion distances
#####

labs_bb <- paste("Dimension", 10:82, "(", 
              round(100*betadisper_bogbean$eig / sum(betadisper_bogbean$eig), 2), "%)")

plot(betadisper_bogbean, cex=1, pch=15:17,
     main="Sampling years (bogbean): MDS coordinates", cex.lab=1.25,
     xlab=labs_bb[1], ylab=labs_bb[2],
     hull=FALSE, ellipse=TRUE, conf=0.68, lwd=2)

boxplot(betadisper_bogbean, xlab="Sampling years (bogbean)", notch=FALSE, col=c("gray", "red", "green"))

```

```{R 1979-2019 diversity calculations}

#######################
# Sedge 
#######################

# select only plots in sedge group
sedge1979 <- d1979_grps %>% 
  filter(Association_Type2 == 1) %>% 
  select(-(Association_Type2:Association_Subtype)) 
#nrow(sedge1979) #34 plots

sedge1999 <- d1999_grps %>% 
  filter(Association_Type2 == 1) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(sedge1999) #31 plots

sedge2019 <- d2019_grps %>% 
  filter(Association_Type2 == 1) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(sedge2019) #25

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#select only species present
sp_present_sedge1979 <- sedge1979[, colSums(sedge1979 != 0) > 0]
sp_present_sedge1999 <- sedge1999[, colSums(sedge1999 != 0) > 0]
sp_present_sedge2019 <- sedge2019[, colSums(sedge2019 != 0) > 0]

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ncol = # species present
ncol(sp_present_sedge1979) #34
ncol(sp_present_sedge1999) #35
ncol(sp_present_sedge2019) #34

##~~~~~~~~~~~~~~~~~~~
# Alpha diversity
##~~~~~~~~~~~~~~~~~~~

sp_plot_sedge1979 <- apply(sp_present_sedge1979 >0,1,sum)
#[1]   7  7  7 10  9  8  8  5  7  9  6  5 10  6 14  8 11  7 11 11 15  8 11  8 10 12 10  8 10 11  6  7 10  5

sp_plot_sedge1999 <- apply(sp_present_sedge1999 >0,1,sum)
#[1]  9  8  8  5 10  7  6 10  8  9  7  7  7  7 11  9  7 10 13  6  8 12  7  8  7  6 10 11 10  8  5

sp_plot_sedge2019 <- apply(sp_present_sedge2019 >0,1,sum)
#[1]  8 11  9 11  9 11  7  2  5  9  7  9 11  6  8 10 10 12  8  7  3  5 10  5  8

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# species density = mean # sp
mean(sp_plot_sedge1979) #8.7
mean(sp_plot_sedge1999) #8.3
mean(sp_plot_sedge2019) #8.0

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#sd of species density
sd(sp_plot_sedge1979) #2.5
sd(sp_plot_sedge1999) #2.0
sd(sp_plot_sedge2019) #2.6

##~~~~~~~~~~~~~~~~~~~
# Beta diversity: total n species divided by density
##~~~~~~~~~~~~~~~~~~~

ncol(sp_present_sedge1979)/mean(sp_plot_sedge1979) #3.9
ncol(sp_present_sedge1999)/mean(sp_plot_sedge1999) #4.2
ncol(sp_present_sedge2019)/mean(sp_plot_sedge2019) #4.2

#######################
# Fescue 
#######################

fescue1979 <- d1979_grps %>% 
  filter(Association_Type2 == 2) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(fescue1979) #29

fescue1999 <- d1999_grps %>% 
  filter(Association_Type2 == 2) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(fescue1999) #33

fescue2019 <- d2019_grps %>% 
  filter(Association_Type2 == 2) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(fescue2019) #14

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#select only species present
sp_present_fescue1979 <- fescue1979[, colSums(fescue1979 != 0) > 0]
sp_present_fescue1999 <- fescue1999[, colSums(fescue1999 != 0) > 0]
sp_present_fescue2019 <- fescue2019[, colSums(fescue2019 != 0) > 0]

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ncol = # species present
ncol(sp_present_fescue1979) #47
ncol(sp_present_fescue1999) #41
ncol(sp_present_fescue2019) #26

##~~~~~~~~~~~~~~~~~~~
# Alpha diversity
##~~~~~~~~~~~~~~~~~~~

sp_plot_fescue1979 <- apply(sp_present_fescue1979 >0,1,sum)
#[1]   7  7  9 15 13 12 12 17 13  8  7 18 17 11 12 16 18 11  8 11 10 12 14 20 20 17 15 11 11
sp_plot_fescue1999 <- apply(sp_present_fescue1999 >0,1,sum)
#[1] 5  6 10  7  9  6  6  5 17 17 14 13 11 16 12 12  7  8 16 12 14  9 10 10 12 15  5  5  6  8  5  5  7

sp_plot_fescue2019 <- apply(sp_present_fescue2019 >0,1,sum)
#[1] 4 10  7  2  6  7 10 11  7  3  5  8  6  7

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# species density = mean # sp
mean(sp_plot_fescue1979) #12.8
mean(sp_plot_fescue1999) #9.7
mean(sp_plot_fescue2019) #6.6

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#sd of species density
sd(sp_plot_fescue1979) #3.9
sd(sp_plot_fescue1999) #3.9
sd(sp_plot_fescue2019) #2.6

##~~~~~~~~~~~~~~~~~~~
# Beta diversity: total n species divided by density
##~~~~~~~~~~~~~~~~~~~

ncol(sp_present_fescue1979)/mean(sp_plot_fescue1979) #3.7
ncol(sp_present_fescue1999)/mean(sp_plot_fescue1999) #4.2
ncol(sp_present_fescue2019)/mean(sp_plot_fescue2019) #3.9

#######################
# Bogbean 
#######################

bogbean1979 <- d1979_grps %>% 
  filter(Association_Type2 == 3) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(bogbean1979) #19

bogbean1999 <- d1999_grps %>% 
  filter(Association_Type2 == 3) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(bogbean1999) #18

bogbean2019 <- d2019_grps %>% 
  filter(Association_Type2 == 3) %>% 
  select(-(Association_Type2:Quadrat)) 
#nrow(bogbean2019) #28

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#select only species present
sp_present_bogbean1979 <- bogbean1979[, colSums(bogbean1979 != 0) > 0]
sp_present_bogbean1999 <- bogbean1999[, colSums(bogbean1999 != 0) > 0]
sp_present_bogbean2019 <- bogbean2019[, colSums(bogbean2019 != 0) > 0]

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ncol = # species present
ncol(sp_present_bogbean1979) #32
ncol(sp_present_bogbean1999) #36
ncol(sp_present_bogbean2019) #34

##~~~~~~~~~~~~~~~~~~~
# Alpha diversity
##~~~~~~~~~~~~~~~~~~~

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
sp_plot_bogbean1979 <- apply(sp_present_bogbean1979 >0,1,sum)
#[1]  12 16 12 11  9 17 12 16 18 19 15 15  6  7 11 14 10 10 14

sp_plot_bogbean1999 <- apply(sp_present_bogbean1999 >0,1,sum)
#[1]  12 14 13 13 14 17 13 10 12 10 12 12 12 10  6  6  7 14

sp_plot_bogbean2019 <- apply(sp_present_bogbean2019 >0,1,sum)
#[1]  10 11  8 11 12 11 11 10 10  7 12 11  9 14 12  9 10  7 11 14 11 12  9 11  6 11 12 11

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# species density = mean # sp
mean(sp_plot_bogbean1979) #12.8
mean(sp_plot_bogbean1999) #11.5
mean(sp_plot_bogbean2019) #10.5

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#sd of species density
sd(sp_plot_bogbean1979) #3.6
sd(sp_plot_bogbean1999) #2.9
sd(sp_plot_bogbean2019) #1.9

##~~~~~~~~~~~~~~~~~~~
# Beta diversity: total n species divided by density
##~~~~~~~~~~~~~~~~~~~

ncol(sp_present_bogbean1979)/mean(sp_plot_bogbean1979) #2.5
ncol(sp_present_bogbean1999)/mean(sp_plot_bogbean1999) #3.1
ncol(sp_present_bogbean2019)/mean(sp_plot_bogbean2019) #3.2
```

```{r invasive species stacked bar}
#https://jkzorz.github.io/2019/06/05/stacked-bar-plots.html
#https://stackoverflow.com/questions/38452577/making-stack-bar-plot-of-bacterial-abundance



join_40_no_outliers$Year <- as.character(join_40_no_outliers$Year)

join_40_no_outliers %>% 
  select(Year, Association_Type2, AGST2:SYSU4) %>% 
  write.csv('join_species.csv', row.names = F)

species <- read.csv("join_species.csv", header = TRUE)
species$Year <- as.character(species$Year)


Ladner_long <- species %>% 
  pivot_longer(AGST2:SYSU4, names_to = "Species", values_to = "Cover") %>% 
  mutate(Species_groups = ifelse(Species %in% c("IRPS", "MYSC", "RUCO2", "CIAR4", "Lycopus_sp", "SOAR2", "POHY", "ALGE2"), "Other Non-native", 
                          ifelse(Species =="AGST2", "Agrostis stolonifera",
                          ifelse(Species == "LYSA2", "Lythrum salicaria",
                          ifelse(Species == "FEAR3", "Festuca arundinaceae",
                          ifelse(Species == "MEAQ", "Mentha aquatica",
                          ifelse(Species == "PHAR3", "Phalaris arundinaceae", 
                          "Native Species")))))))
 Ladner_long$Species_groups <- factor(Ladner_long$Species_groups, 
                                      levels = c("Agrostis stolonifera", "Festuca arundinaceae", 
                                                 "Lythrum salicaria", "Mentha aquatica", 
                                                 "Phalaris arundinaceae","Other Non-native","Native Species"))

# what would this be, avg. relative abundance per plot? How is Abundance calculated??
 library(RColorBrewer)
 library(ggtext)

 italics <- expression(italic("Agrostis stolonifera"), 
                       italic("Festuca arundinaceae"), 
                       italic("Lythrum salicaria"), 
                       italic("Mentha aquatica"), 
                       italic("Phalaris arundinaceae"),"Other Non-native","Native Species")
 
ggplot(Ladner_long, aes(fill=Species_groups, y=Cover, x=Year)) + 
    geom_bar(position="fill", stat="identity") + theme_classic()+
    facet_grid(~Association_Type2, scales = "free", space = "free") + 
    theme(text = element_text(size = 16)) +
    theme(panel.spacing.x = unit(1.25, "lines")) +
    theme(axis.text.x = element_text(angle = 45, margin = margin(t=20, r=100)))+
    scale_fill_brewer(name = "", palette = "Set3", labels = italics) +
    theme(legend.text.align = 0) +
    labs(y="Relative Abundance 
         (proportion of plot cover)", x="Observation Date") 
#ggsave("Invasive.svg")
#ggsave("Invasive.png")


# Ladner_abund <- Ladner_long %>% 
#   group_by(Year, Association_Type2, Endemism) %>% 
#   dplyr::summarise(
#     n = n(),
#     mean = mean(Cover, na.rm = TRUE),
#     sd = sd(Cover, na.rm = TRUE),
#     se = sd/sqrt(n))


#~~~~~~~~~~~~~ see also
#https://kembellab.ca/r-workshop/biodivR/SK_Biodiversity_R.html

```

```{r community proximity to channel}

ggplot(data=d2019_grps, 
    aes(x= Association_Type2, y=Dist_channel_m, group = Association_Type2)) +
    theme(text = element_text(size = 20)) +
    geom_boxplot(size=1, alpha = 1/10)+ theme_classic() +
  labs(y="Distance from nearest channel edge (m)", x="Community Type") +
  scale_x_continuous(breaks = 1:3, labels = c("Sedge", "Fescue", "Bogbean"))

####################################
# ANOVA of distance to channel
####################################
library(rstatix)
library(ggpubr)

d2019_grps$Association_Type2 <- as.factor(d2019_grps$Association_Type2)

# build model, check assumptions
comm_model <- lm(Dist_channel_m ~ Association_Type2, data = d2019_grps)
ggqqplot(residuals(comm_model))
plot(comm_model, 1)

ANOVA_comm <- d2019_grps %>% anova_test(Dist_channel_m ~ Association_Type2)
ANOVA_comm

tukey_comm <- d2019_grps %>% tukey_hsd(Dist_channel_m ~ Association_Type2)
tukey_comm # sedge and bogbean are significantly different distances from channel edge (p = 0.0045), but not sedge:fescue or fescue:bogbean
# could suggest that fesuce-type is encroaching inward into marsh, as invasive species tolerant of saturation (RCG, bentgrass, loosestrife) can make inroads

```

```{r 2019 PCA of community types}
###########################
# PCA
###########################
#from  OurCodingClub

#eigenvalues
PCA2019 <- rda(d2019_select, scale = FALSE)

# Now plot a bar plot of relative eigenvalues. This is the percentage variance explained by each axis
barplot(as.vector(PCA2019$CA$eig)/sum(PCA2019$CA$eig))
# How much of the variance in our dataset is explained by the first principal component? -> > 25%


# Calculate the percent of variance explained by first two axes
sum((as.vector(PCA2019$CA$eig)/sum(PCA2019$CA$eig))[1:2]) #42% of variance explained by first two axes

sum((as.vector(PCA2019$CA$eig)/sum(PCA2019$CA$eig))[1]) #28.7% of variance explained by first axis
sum((as.vector(PCA2019$CA$eig)/sum(PCA2019$CA$eig))[2]) #13.3% of variance explained by second axis

plot(PCA2019)
plot(PCA2019, display = "sites", type = "text")
plot(PCA2019, display = "species", type = "text")

# You can extract the species and site scores on the new PC for further analyses:
sitePCAmed <- PCA2019$CA$u # Site scores
speciesPCAmed <- PCA2019$CA$v # Species scores

# In a biplot of a PCA, species' scores are drawn as arrows 
# that point in the direction of increasing values for that variable
biplot(PCA2019, 
       choices = c(1,2), 
       cex = c(5, 5, 5),
       xlab = "PC1 (28.7%)", 
       ylab = "PC2 (13.3%)")


  #main = "2019 Plant Community PCA" # biplot of axis 1 vs 2

```

```{r unused/reference code -> DEL when done, echo=FALSE}

#####################################
# Trying to recreate Table 2 (B-diversity) from B&P per Methods
#####################################
# 1. calculate alpha diversity (species density = mean # sp per quadrat)
# 2. beta diversity is a ratio of total # species divided by species density

#all 1979 data
d1979

# only species data for community 1a 
IA1979 <- d1979 %>% 
  filter(Association_Subtype == "1a") %>% 
  select(-(Year:Association_Subtype)) 

nrow(IA1979) #15

sp_present <- IA1979[, colSums(IA1979 != 0) > 0]

#####
# Alpha diversity
#####

# (#) sp per quadrat = sum up the number of non-zero entries per row (1)
sp_quad <- apply(IA1979 >0,1,sum)
#  [1]  6  8 11 10 12 10  8 12 12 11 10 11  6  7 10

sp_1A <- apply(sp_present >0,1, sum)

ncol(sp_present)
colnames(sp_present) #29
#  [1] "AGST2"  "ALPL"   "BICE"   "CALY3"  "CAPAP6" "DECE"   "DUAR3"  "ELPA3"  "EQFL"   "FEAR3"  "IMCA"   "LAPA4" 
# [13] "LEOR"   "LIOC"   "LYSA2"  "MEAR4"  "MIGU"   "OESA"   "PLDID"  "POPA2"  "POPA23" "PUPA3"  "RUOC3"  "SALA2" 
# [25] "SASI2"  "SCTA2"  "SISU2"  "SYEA2"  "TYLA"  


# species density = mean # sp
mean(sp_quad) #9.6


#####
# Beta diversity: total n species dividied by density
#####

ncol(sp_present)/mean(sp_quad) #3


#######################
# Analysis of Similarity
#######################
#note: ANOSIM is sensitive to inequal compositional variance (see Anderson & Walsh, 2013)
# these attempts use full years' dataset, not by community

#analysis of similarity between years of observation
ano_out = anosim(m_com_out, join_40_no_outliers$Year, distance = "bray", permutations = 9999)
ano_out
# ANOSIM statistic R: 0.2021
#       Significance: 1e-04 

#analysis of similarity between community types, combined across all years
ano_assoc_out = anosim(m_com_out, join_40_no_outliers$Association_Type2, distance = "bray", permutations = 9999)
ano_assoc_out

# ANOSIM statistic R: 0.351 
#       Significance: 1e-04 

#######################
# PERMANOVA
#######################
permanova_out = adonis(m_com_out ~ join_40_no_outliers$Year, permutations = 999, method = "bray")


#                           Df SumsOfSqs MeanSqs F.Model      R2 Pr(>F)    
# join_40_no_outliers$Year   1     5.828  5.8282  23.221 0.09028  0.001 ***
# Residuals                234    58.731  0.2510         0.90972           
# Total                    235    64.560                 1.00000 

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# Diversity indices
#from http://benbestphd.com/landscape-ecology-labs/lab8.html#other-diversity-indices
# see also section 4 from https://www.flutterbys.com.au/stats/tut/tut13.2.html

# Generates index per plot - how to demonstrate for entire community?
Shannon2019 = diversity(d2019_select, index='shannon')
Simpson2019 = diversity(d2019_select, index='simpson')
InvSimpson2019 = diversity(d2019_select, 'inv')
Rare2019 = rarefy(round(d2019_select), 2) - 1
Alpha2019 = fisher.alpha(round(d2019_select))

pairs(cbind(Shannon2019, Simpson2019, InvSimpson2019, Rare2019, Alpha2019), pch='+', col='blue')

plot(Simpson2019)


######################################################################
#1979-2019 NMDS comparison (outliers removed);

join_40 <- lst(d1979_grps, d1999_grps, d2019_grps) %>% reduce(full_join) %>% replace(is.na(.), 0)
nrow(join_40) #238

#After running NMDS, found outliers (code containing outliers in next chunk). 
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# two outliers: 2019 sites W_29 and X_43 are outliers on NMDS plot. remove here
join_40_no_outliers <- join_40[!(join_40$Year == 2019 & join_40$Quadrat == 29) & !(join_40$Year == 2019 & join_40$Quadrat == 43),]
#nrow(join_40_no_outliers) #236
  

#head(join_40_no_outliers) #sp data start in 7th col

#rerun NMDS w/ outliers removed. All descriptive text is removed here. 
com_out = join_40_no_outliers[,7:ncol(join_40_no_outliers)]
m_com_out = as.matrix(com_out)
set.seed(123)
nmds_out = metaMDS(m_com_out, distance = "bray")
nmds_out

data.scores_out = as.data.frame(scores(nmds_out))

data.scores_out$Year = join_40_no_outliers$Year
data.scores_out$Association_Type2 = join_40_no_outliers$Association_Type2

Ladner.scores_out <- data.scores_out %>% 
  mutate(Year = as.character(Year)) %>% 
           mutate(Association_Type2 = as.character(Association_Type2))

Ladner40yrs_out = ggplot(Ladner.scores_out, aes(x = NMDS1, y = NMDS2)) + 
    geom_point(size = 5, aes(shape = Association_Type2, colour = Year))+ 
    theme(axis.text.y = element_text(colour = "black", size = 16, face = "bold"), 
    axis.text.x = element_text(colour = "black", face = "bold", size = 16), 
    legend.text = element_text(size = 14, face ="bold", colour ="black"), 
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
    axis.title.x = element_text(face = "bold", size = 16, colour = "black"), 
    legend.title = element_text(size = 14, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank()) + 
    labs(x = "NMDS1", colour = "Year", y = "NMDS2", shape = "Community Association Type")  + 
    scale_colour_manual(values = c("#4B0082", "#9370DB", "#DDA0DD"))   #+
    # geom_path(data=df_ell, aes(x=NMDS1, y=NMDS2,colour=Year), size=1, linetype=2) #+
    # #annotate("text",x=NMDS.mean$MDS1,y=NMDS.mean$MDS2,label=NMDS.mean$Year)
    # 
Ladner40yrs_out

```

``` {r maps, echo=FALSE}
# https://geocompr.robinlovelace.net/adv-map.html#static-maps
# 
# install.packages("sf")
# install.packages("raster")
# install.packages("spData")
# install.packages("spDataLarge")
# install.packages("tmap")
library(sf)
library(raster)
library(spData)
library(spDataLarge)
library(tmap)

```
